---
title: "Getting Started with dsOMOP v2"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with dsOMOP v2}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(eval = FALSE)
```

## Overview

**dsOMOP v2** provides safe, federated access to OMOP Common Data Model databases
through the DataSHIELD framework. It consists of two packages:

- **dsOMOP** (server-side): runs on each DataSHIELD node. Handles database
  queries, disclosure control, and data preparation.
- **dsOMOPClient** (client-side): runs on your local machine. Provides an
  ergonomic API for schema exploration, cohort management, and data extraction.

## Prerequisites

- A DataSHIELD server (Opal, Armadillo, or DSLite) with dsOMOP v2 installed
- An OMOP CDM database configured as a DataSHIELD resource
- The `dsOMOPClient` package installed locally

```{r install}
remotes::install_github("isglobal-brge/dsOMOPClient")
```

## Quick Start

### 1. Connect to DataSHIELD

```{r connect}
library(dsOMOPClient)
library(DSI)
library(DSOpal)

builder <- DSI::newDSLoginBuilder()
builder$append(
  server = "server1",
  url = "https://opal.example.org",
  user = "analyst",
  password = "secret",
  resource = "project.omop_cdm"
)

login_data <- builder$build()
conns <- DSI::datashield.login(logins = login_data)
```

### 2. Attach the OMOP Resource

```{r attach}
ds.omop.connect(
  resource = "project.omop_cdm",
  conns = conns,
  symbol = "omop"
)
```

This initializes a server-side handle that connects to the database, introspects
the schema, and prepares for queries. The session is stored under `symbol`
(default `"omop"`). Every later call defaults to `symbol = "omop"`, so in a
single-session workflow you can omit it entirely.

### 3. Explore the Schema

```{r explore}
ds.omop.tables(symbol = "omop", conns = conns)            # available tables
ds.omop.columns("person", symbol = "omop", conns = conns)  # columns of a table
ds.omop.table.stats("person", symbol = "omop", conns = conns)  # safe row/person counts
```

See the *Schema Exploration* vignette for prevalence, statistics, missingness,
and the disclosure-gated manipulation verbs.

### 4. Extract Data with a Plan

The plan DSL specifies exactly what to extract:

```{r plan}
plan <- ds.omop.plan()
plan <- ds.omop.plan.baseline(plan,
  columns = c("gender_concept_id", "year_of_birth"))
plan <- ds.omop.plan.events(plan,
  name = "conditions",
  table = "condition_occurrence",
  concept_set = c(201826, 255573))

print(plan)
```

### 5. Preview and Execute

`ds.omop.plan.preview()` is a server-side dry run: it returns each output's
expected columns and a banded person-count estimate **without creating any
data**.

```{r preview}
ds.omop.plan.preview(plan, symbol = "omop", conns = conns)
```

`ds.omop.plan.execute()` assigns each output into the DataSHIELD session. The
`out` argument is a **named** vector mapping plan output names to server-side
symbols:

```{r execute}
ds.omop.plan.execute(plan,
  out = c(baseline = "D", conditions = "E"),
  symbol = "omop",
  conns = conns
)

# D and E are now server-side data frames
dsBaseClient::ds.summary("D")
dsBaseClient::ds.dim("E")
```

### 6. Clean Up

```{r cleanup}
ds.omop.disconnect(symbol = "omop", conns = conns)
DSI::datashield.logout(conns)
```

## Ergonomics

### Recipes: named, reusable, save/loadable extractions

For anything beyond a one-off plan, use the **recipe** API. A recipe is a named
container of variables and outputs that compiles to a plan, and it can be saved
to and loaded from disk:

```{r recipe}
rec <- omop_recipe()
rec <- recipe_add_variable(rec,
  omop_variable(table = "measurement", concept_id = 3004410, format = "mean"))
rec <- recipe_add_output(rec, omop_output(name = "study", type = "features"))

# With no `out`, recipe_execute() derives the output symbols from the output
# names (here "study" -> "D_study") -- the simplest single-output form:
recipe_execute(rec, symbol = "omop", conns = conns)

# Or assign explicit symbols, exactly like the plan API:
recipe_execute(rec, out = c(study = "D"), symbol = "omop", conns = conns)

# Persist the recipe (format inferred from the extension; YAML or JSON):
recipe_save(rec, "study.yaml")
rec <- recipe_load("study.yaml")
```

See the *Data Extraction* vignette for the full recipe and plan reference,
including the row-level filter DSL, per-variable time windows, unit/type concept
scoping, and visit linkage.

## Where to Go Next

- **Schema Exploration** -- prevalence, statistics, and the manipulation verbs.
- **Cohort Management** -- spec-based, workspace, and pre-existing cohorts; the
  unified `cohort =` argument.
- **Data Extraction** -- the plan DSL, recipes, and max-flexibility extraction.
- **Multi-Server** -- federated analysis and plan harmonization.
- **Security** -- the per-patient gate, the person token, and admin configuration.
