---
title: "1. Creating an interface object"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{01_object}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

**Important note:** The `dsOMOPClient` package serves as a client-side interface with DataSHIELD servers, facilitating the retrieval of tables from OMOP CDM databases based on resources and integrating such tables into the DataSHIELD workflow. However, users will need to manually carry out the subsequent processes of transforming the resulting tables into a single dataset afterwards using the DataSHIELD base functions of `dsBaseClient`. 

There are auxiliary packages that automate this integration, which is why we recommend taking a look at packages like `dsOMOPHelper` instead of directly using `dsOMOPClient`. We advise against the direct use of `dsOMOPClient` for regular researchers unless faced with an edge-case usage that requires the flexibility of operating directly with the `dsOMOPClient` data interface at a more primitive level, or for developers who are building their own packages on top of `dsOMOPClient`. 

For more information on `dsOMOPHelper`, please visit [its GitHub repository](https://github.com/isglobal-brge/dsOMOPHelper).

# Prerequisites

Before using `dsOMOPClient`, it is recommended to have a basic understanding of:

- The OMOP CDM structure and its standardized clinical data format. You can learn more about OMOP CDM in the [OHDSI Book chapter 'The Common Data Model'](https://ohdsi.github.io/TheBookOfOhdsi/CommonDataModel.html).

- OMOP Vocabularies and how they standardize medical concepts (like diagnoses, medications, procedures) across different coding systems (ICD-9, ICD-10, SNOMED CT, etc.) into a common representation. The [OHDSI Book chapter 'Standardized Vocabularies'](https://ohdsi.github.io/TheBookOfOhdsi/StandardizedVocabularies.html) provides a comprehensive overview of this standardization process.

- Basic DataSHIELD concepts and workflow. The [DataSHIELD Beginner's Tutorial](https://data2knowledge.atlassian.net/wiki/spaces/DSDEV/pages/931069953/Beginners+Tutorial+DataSHIELD+v6.1) is a good starting point.

This knowledge will help you better understand how to effectively query and work with OMOP CDM data through the DataSHIELD infrastructure.

# 1.1 Creating an interface object

The `ds.omop` function creates an interface object that allows users to interact with the OMOP CDM database based on a resource. We can use this object to obtain tables from the database by applying the desired filters and querying data catalogs for information present in the database. 

In this example, we will be using the [MIMIC IV](https://physionet.org/content/mimic-iv-demo-omop/0.9/) data available on the [OBiBa's public Opal demo server](https://opal-demo.obiba.org). This server is publicly accessible, so all users are able to reproduce the examples of this guide by executing the same commands in their R session. The access credentials are:

- Server URL: `https://opal-demo.obiba.org`
- User: `dsuser`
- Password: `P@ssw0rd`
- Profile: `omop`

First, we will establish a connection to the demo server using `DSI` with the provided credentials:

```{r setup, include=FALSE}
library(knitr)
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE
)
```

```{r establishing_connection, message=FALSE}
library(DSI)
library(DSOpal)
library(dsBaseClient)
library(dsOMOPClient)

builder <- newDSLoginBuilder()
builder$append(
  server = "opal_demo",
  url = "https://opal-demo.obiba.org",
  user = "dsuser",
  password = "P@ssw0rd",
  profile = "omop"
)
logindata <- builder$build()
conns <- datashield.login(logins = logindata)
```

Once we have successfully established a connection with the server, we will create the object with `ds.omop`. The function requires the following parameters:

- `connections`: A list of established DataSHIELD connections.
- `resource`: The name of the resource of the OMOP CDM database in the DataSHIELD server.

Our server contains the database connection resource under the name `mimiciv` within the `omop_demo` project. Therefore, we need to specify that, from the connection we have established, we want to take the `omop_demo.mimiciv` resource.

```{r creating_object}
o <- ds.omop(
  connections = conns,
  resource = "omop_demo.mimiciv"
)
```

## 1.1.1 Establishing multiple connections

If we wish to establish connections with multiple OMOP CDM databases simultaneously in a `connections` pool, the `resource` parameter accepts a named list as a parameter where the name of the list object corresponds to the server name, and the value corresponds to the resource name. For example:

```{r multiple_connections, eval = FALSE}
o <- ds.omop(
  connections = conns,
  resource = list(opal_demo = "omop_demo.mimiciv",
                  another_server = "project_name.resource_name")
)
```

# 1.2 Interface object functions

The `ds.omop` interface object provides several functions to interact with the OMOP CDM database. These functions allow users to query information from the database and extract tables, which will then be transformed and integrated into the DataSHIELD workflow.

## 1.2.1 Querying information

- `tables()`: This function lists all the available tables in the OMOP CDM database. It allows users to understand the structure of the database and identify which tables might be relevant for their analysis.

```{r tables_catalog}
o$tables()
```

- `columns(tableName)`: Given a table name, this function returns the columns available in that table. This is useful for users to identify the specific data fields they might want to analyze or use in their queries.

```{r columns_catalog}
o$columns("measurement")
```

- `concepts(tableName)`: This function retrieves the concepts present in a given table and returns them as a data frame with two columns: `concept_id` and their associated `concept_name`. Concepts in OMOP CDM are standardized terms that represent clinical events, measurements, observations, and other entities. Understanding the concepts available in a table can help users to formulate data filtering queries.

```{r concepts_catalog, R.options = list(max.print = 20)}
o$concepts("measurement")
```

The numeric values in the `concept_id` column are the standardized identifiers for the concepts in the OMOP CDM. For example, as we can observe in the table above, the concept `1175625` refers to the concept `Breath rate spontaneous`. We will use these identifiers to filter the data in the following sections.

## 1.2.2 Retrieving tables

- `get(tableName)`: This function enables users to extract a specific table from the OMOP CDM database. The extracted table becomes available within the DataSHIELD environment, where it can be manipulated using other DataSHIELD functions and potentially combined with other tables from the database. Users have the flexibility to apply filters and specify columns to customize the extracted data according to their research needs. Below are some examples of how to use the `get` function:

### Getting a complete table

We simply specify the desired table's name as a string in the `get` function: 

```{r get_person}
o$get("person")

# We can use the `ds.summary` function to get a summary of the retrieved table
ds.summary("person")
```

However, this is only advised for small tables or in contexts where we need the full table, as it can be memory-intensive and slow to retrieve. For larger tables, it is recommended to apply filters to the `get` function to retrieve a specific subset of the table.

### Getting a filtered table

We can apply filters to the `get` function to retrieve a specific subset of the table, which is the recommended approach for larger tables, such as the `Measurement` table. For example, we can set the parameters of the `get` function to retrieve only specific columns (in this case, `value_as_number` and `measurement_date`) and specific measurement types (in this case, we want to use the concept IDs of `Heart rate` and `Body weight`, which are `3027018` and `3025315`):

```{r get_measurement, R.options = list(max.print = 5)}
o$get(table = "measurement",
      columnFilter = c("value_as_number", "measurement_date"),
      conceptFilter = c(3027018, 3025315))

ds.summary("measurement")
```

As we can observe, the resulting table contains only the information about the measurements of `Heart rate` and `Body weight` of all patients in the database.

These functions form the core of the `ds.omop` interface object's capabilities, which enables users to interact with OMOP CDM databases within the DataSHIELD environment. It is expected that the various resulting tables will be relationally joined using the `ds.merge` function through their common identifiers, such as `person_id`.

In the next article, we will explore the internal workings of the table processing operations, understanding how to manipulate and combine tables into single datasets ready for analysis, along with a deeper dive into the customization possibilities of the `get` function and its potential applications.

```{r logout, echo = FALSE}
datashield.logout(conns)
```
