---
title: "Schema Exploration"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Schema Exploration}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(eval = FALSE)
```

## Overview

Before extracting data, you need to understand what's available in each CDM.
dsOMOP v2 provides a set of **aggregate** exploration functions that return safe
metadata without exposing individual-level data.

Every exploration function respects DataSHIELD disclosure controls. Counts that
describe too few unique persons are suppressed or blocked (see the *Security*
vignette). Each function returns a `dsomop_result` object with a `$per_site`
list (one entry per server), an optional `$pooled` slot (when `scope = "pooled"`),
and a `$meta` slot that records the generated call and any server warnings.

## Prerequisites

```{r connect}
library(dsOMOPClient)
# Assume connections are established and the OMOP resource is attached:
# ds.omop.connect(resource = "project.omop_cdm", conns = conns, symbol = "omop")
```

## Listing Tables and Columns

```{r tables}
tables <- ds.omop.tables(symbol = "omop", conns = conns)
# Per-site character vectors of discovered table names

cols <- ds.omop.columns("condition_occurrence", symbol = "omop", conns = conns)
# Per-site column metadata: name, type, nullability
```

dsOMOP discovers tables and columns at runtime through schema introspection. It
does not hard-code OMOP CDM version tables, so custom tables appear here too.

## Relationships

```{r joins}
joins <- ds.omop.joins(symbol = "omop", conns = conns)
# Inferred join edges (person_id, visit_occurrence_id, *_concept_id, ...)
```

## Table and Column Statistics

```{r stats}
ts <- ds.omop.table.stats("measurement", symbol = "omop", conns = conns)
# row_count and person_count (COUNT(DISTINCT person_id)); suppressed if small

cs <- ds.omop.column.stats("measurement", "value_as_number",
  concept_id = 3004410, symbol = "omop", conns = conns)
cs$per_site$server1
```

`ds.omop.column.stats()` returns a disclosure-safe numeric summary: `n`, `mean`,
**`sd`** (standard deviation), `min`, `max`, and quantiles. The standard
deviation is computed server-side and returned alongside the mean so you can
characterise spread without pulling any row-level values.

## Concept Prevalence

The most frequent concepts in a table, ranked by distinct persons (default) or
records:

```{r prevalence}
prev <- ds.omop.concept.prevalence("condition_occurrence",
  metric = "persons", top_n = 50, symbol = "omop", conns = conns)
head(prev$per_site$server1)
```

### Person-column prevalence (demographics)

Pass an explicit `concept_col` to rank the categorical concepts on the `person`
table. This is how you obtain gender, race, and ethnicity breakdowns:

```{r person-cols}
ds.omop.concept.prevalence("person", concept_col = "gender_concept_id",
  symbol = "omop", conns = conns)
ds.omop.concept.prevalence("person", concept_col = "race_concept_id",
  symbol = "omop", conns = conns)
ds.omop.concept.prevalence("person", concept_col = "ethnicity_concept_id",
  symbol = "omop", conns = conns)
```

### Death table

The `death` table is explored the same way, naming its cause concept column:

```{r death}
ds.omop.concept.prevalence("death", concept_col = "cause_concept_id",
  symbol = "omop", conns = conns)
```

### All-concepts / global prevalence with pagination

`global = TRUE` ranks concepts across **all** person-bearing CDM tables in one
merged, re-gated ranking instead of a single table. Use `offset` together with
`top_n` to page through the ranking (the page is `[offset+1 .. offset+top_n]`):

```{r global}
# First page of the 25 most prevalent concepts across the whole CDM
page1 <- ds.omop.concept.prevalence(global = TRUE, top_n = 25,
  symbol = "omop", conns = conns)

# Next page
page2 <- ds.omop.concept.prevalence(global = TRUE, top_n = 25, offset = 25,
  symbol = "omop", conns = conns)
```

Small cells are suppressed over the merged set *before* the page is taken, so
pagination never reveals a count that the single-table gate would have blocked.

## Value Frequencies and Histograms

```{r valuecounts}
ds.omop.value.counts("person", "gender_concept_id",
  symbol = "omop", conns = conns)

ds.omop.value.histogram("measurement", "value_as_number",
  concept_id = 3004410, symbol = "omop", conns = conns)

ds.omop.value.quantiles("measurement", "value_as_number",
  symbol = "omop", conns = conns)
```

Value counts enforce small-cell suppression and a maximum number of distinct
levels (see *Security*).

## Other Profiling

```{r misc}
ds.omop.domain.coverage(scope = "pooled", symbol = "omop", conns = conns)
ds.omop.missingness("measurement",
  columns = c("value_as_number", "value_as_concept_id"),
  symbol = "omop", conns = conns)
ds.omop.crosstab("person", "gender_concept_id", "race_concept_id",
  symbol = "omop", conns = conns)
```

## Scoping Exploration to a Cohort

Most exploration functions accept a unified `cohort =` argument so you can run
the same profiling within a subpopulation. The argument accepts a cohort handle
(from `ds.omop.cohort.create()` / `ds.omop.cohort.combine()` /
`ds.omop.cohort.from_table()`), a numeric cohort definition id, or a server-side
cohort table name:

```{r cohort-scope}
dm <- ds.omop.cohort.create(
  spec = list(type = "condition", concept_set = c(201826)),
  symbol = "omop", conns = conns)

# Prevalence, statistics and missingness restricted to the diabetes cohort
ds.omop.concept.prevalence("drug_exposure", cohort = dm,
  symbol = "omop", conns = conns)
ds.omop.column.stats("measurement", "value_as_number",
  concept_id = 3004410, cohort = dm, symbol = "omop", conns = conns)
ds.omop.missingness("measurement", cohort = dm,
  symbol = "omop", conns = conns)
```

See the *Cohort Management* vignette for building cohorts, including turning a
workspace `omop.table` into a reusable cohort with
`ds.omop.cohort.from_table()`.

## Disclosure-Gated Manipulation of Extracted Tables

Once a plan or recipe has produced a server-side `omop.table` (see *Extraction*),
you can reshape it **without** pulling any row-level data using four
disclosure-aware verbs. Each verb takes the *name(s)* of server-side `omop.table`
symbols, computes the new frame on every server, and stores it under `newobj`.
Only `omop.table` inputs are accepted; the result is re-gated on the number of
**distinct person tokens** and re-tagged as an `omop.table`, so a verb can only
ever narrow a population and fails closed (errors) when the result would describe
too few individuals.

```{r verbs}
# Join two token-keyed frames on the person key (only the person key is allowed)
joined <- ds.omop.merge("baseline", "labs", by = "person_id", type = "left",
  symbol = "omop", conns = conns)

# Keep a subset of rows with a single comparison (the person token is protected)
adults <- ds.omop.filter("baseline", var = "age", op = ">=", value = 18,
  symbol = "omop", conns = conns)

# Project to a subset of columns (the person token is always retained)
slim <- ds.omop.select(joined, cols = c("age", "gender_concept_id"),
  symbol = "omop", conns = conns)

# Row-bind two schema-identical frames; the gate counts DISTINCT persons,
# so binding a frame to itself cannot inflate the count past the threshold
stacked <- ds.omop.bind_rows("wave1", "wave2", symbol = "omop", conns = conns)
```

Each call returns (invisibly) the name of the new server-side symbol, which you
can feed into further verbs, into `ds.omop.cohort.from_table()`, or into
`dsBaseClient` analysis functions.

## Multi-Server Schema Comparison

When working with several servers, compare their schemas before planning:

```{r compare}
comparison <- ds.omop.compare(symbol = "omop", conns = conns)
comparison$common_tables   # tables present on every server
comparison$server_only     # tables unique to particular servers
comparison$column_diffs    # column differences across servers
```

This comparison feeds plan harmonization (see the *Multi-Server* vignette).

## Schema Snapshot

```{r snapshot}
snapshot <- ds.omop.snapshot(symbol = "omop", conns = conns)
# Tables, columns and relationships per server, for documentation/visualisation
```
