---
title: "2. Retrieving tables from the database"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{02_tables}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r creating_object, include = FALSE}
library(DSI)
library(DSOpal)
library(dsBaseClient)
library(dsOMOPClient)

builder <- newDSLoginBuilder()
builder$append(
  server = "opal_demo",
  url = "https://opal-demo.obiba.org",
  user = "dsuser",
  password = "P@ssw0rd",
  profile = "omop"
)
logindata <- builder$build()
conns <- datashield.login(logins = logindata)

o <- ds.omop(
  connections = conns,
  resource = "omop_demo.mimiciv"
)
```

We have already seen how to create an interface object for an OMOP CDM database. Now, let's see how to use it to retrieve tables from it in deeper detail. For this purpose, we will use the `get` method of the interface object. This method has several arguments that allow filtering the data that is retrieved and customizing the output.

# 2.1 Features of the `get` method

##  2.1.1 Table reshaping

The `get` method is an smart method that returns the table after subjecting it to a series of operations. If it detects that this table can present various records linked to the same OMOP CDM entity (for example, to a `Person`), it will transform the table into a wide format automatically using the in-built DataSHIELD reshape function, ready to be merged with other tables. For instance, when we take the `Person` table, it will return a table without altering its original format:

```{r get_person}
o$get("person")

ds.summary("person")
```

Whereas if we take the `Measurement` table, which has records potentially linkable to the `Person` table, it will return a table in wide format, preserving the column `person_id` so that it can be merged with the `Person` table:

```{r get_measurement}
# 3027018 is the concept ID for `Heart rate`
o$get(table = "measurement", conceptFilter = 3027018)

ds.summary("measurement")
```

## 2.1.2 Concept translation

Another feature you may have noticed in the previous example is that the `get` method automatically translates the concepts of the table that we are retrieving. For instance, in the case of the `Measurement` table, the concept `3027018` was translated to `Heart rate`.

The same happens with the values of the rows in the tables. For example, in the `Person` table, in the `gender_concept_id` column, the value `8532` was translated automatically to `FEMALE`, and the value `8507` was translated to `MALE`:

```{r concept_translation}
ds.summary("person$gender_concept_id")
```

This will happen with all concepts in the database as long as they are recognized and properly named in the `Concept` table of the database. Otherwise, the concept will remain as `concept_id_` + the numeric value of the concept. For example, if the concept `FEMALE` was not properly named in the `Concept` table, it would remain as `concept_id_8532` so that it can at least be identified.

## 2.1.3 Longitudinal data sequencing

When you have multiple records for the same entity (e.g., a patient) across time points, `dsOMOP` offers two parameters to transform and align these data in a wide format:

1. **`wideLongitudinal = TRUE`** reshapes repeated measurements for each concept into extra columns, so each row in your table corresponds to one entity (e.g., one patient).
2. **`completeTimePoints = TRUE`** further ensures that all entities share the same set of time points. Rows (and columns) for missing measurements at particular dates are filled with `NA`.

```{r wide_format, R.options = list(max.print = 10)}
o$get(
  table = "measurement",
  conceptFilter = 3027018,
  columnFilter = c("measurement_date", "value_as_number"),
  wideLongitudinal = TRUE,
  completeTimePoints = TRUE
)

ds.summary("measurement")
```

Because DataSHIELD does not allow row-level inspection on the client side, you will not see the exact final structure of the data. Internally, however, `dsOMOP` arranges your dataset so that every individual is aligned on a consistent time axis. Below is a simplified illustration of how this reshaping would look for a small example of three patients (A, B, C). It does *not* reflect real data—merely a conceptual snapshot.

---

**Case 1: `wideLongitudinal = TRUE` but `completeTimePoints = FALSE`**

In this scenario, `dsOMOP` just generates a wide table indexed by the *chronological order* of each patient’s events. It does *not* force every patient to share the same time points. Suppose the patients have data as follows:

- Patient A has 3 heart-rate measurements (Dates: 2023-01-01, 2023-01-15, 2023-01-31)
- Patient B has 2 measurements (Dates: 2023-01-01, 2023-01-02)
- Patient C has 1 measurement (Date: 2023-01-10)

The resulting table, conceptually, might look like this:

| person_id | heart_rate.measurement_date.1 | heart_rate.measurement_value.1 | heart_rate.measurement_date.2 | heart_rate.measurement_value.2 | heart_rate.measurement_date.3 | heart_rate.measurement_value.3 |
|-----------|--------------------------------|--------------------------------|--------------------------------|--------------------------------|--------------------------------|--------------------------------|
| A         | 2023-01-01                    | 70                             | 2023-01-15                    | 71                             | 2023-01-31                    | 72                             |
| B         | 2023-01-01                    | 60                             | 2023-01-02                    | 65                             | NA                             | NA                             |
| C         | 2023-01-10                    | 67                             | NA                             | NA                             | NA                             | NA                             |

- Columns “.1”, “.2”, “.3” represent *the 1st, 2nd, and 3rd measurement* recorded for each patient, in that patient’s *own* chronological order.
- If a patient doesn’t have a second or third measurement, those columns are `NA`.
- Different patients can have entirely different date values in, say, `.2` or `.3`.

---

**Case 2: `wideLongitudinal = TRUE` and `completeTimePoints = TRUE`**

Now `dsOMOP` additionally ensures that *all* patients align on *every* time point found across all patients in the dataset. For instance, if the union of dates in your dataset is `{2023-01-01, 2023-01-02, 2023-01-10, 2023-01-15, 2023-01-31}`, then each row (i.e., each patient) will have columns for *five* distinct measurements, even if some are `NA`:

| person_id | heart_rate.measurement_date.1 | heart_rate.measurement_value.1 | heart_rate.measurement_date.2 | heart_rate.measurement_value.2 | heart_rate.measurement_date.3 | heart_rate.measurement_value.3 | heart_rate.measurement_date.4 | heart_rate.measurement_value.4 | heart_rate.measurement_date.5 | heart_rate.measurement_value.5 |
|-----------|--------------------------------|--------------------------------|--------------------------------|--------------------------------|--------------------------------|--------------------------------|--------------------------------|--------------------------------|--------------------------------|--------------------------------|
| A         | 2023-01-01                    | 70                             | NA                             | NA                             | 2023-01-15                    | 71                             | NA                             | NA                             | 2023-01-31                    | 72                             |
| B         | 2023-01-01                    | 60                             | 2023-01-02                    | 65                             | NA                             | NA                             | NA                             | NA                             | NA                             | NA                             |
| C         | NA                             | NA                             | NA                             | NA                             | 2023-01-10                    | 67                             | NA                             | NA                             | NA                             | NA                             |

- For privacy reasons, it *still* uses index-based columns (e.g., `.1, .2, .3`), but now behind the scenes, each suffix corresponds to the *same date* for every patient.
- Patients missing a measurement on a particular date have `NA` in both the `measurement_date` and `measurement_value` columns for that slot.
- This approach ensures a consistent date axis across all individuals, which is typically needed for time-series analysis or for creating time-aligned comparisons.

---

#### Practical usage notes

1. **Date sorting**: Under the hood, **dsOMOP** first discovers all unique dates in the data, sorts them in ascending order, and then indexes them as `.1, .2, .3, …`.
2. **Flexibility**: You can still opt *not* to align time points by leaving `completeTimePoints` as `FALSE`. This can be useful for scenarios where each patient’s internal timeline is more relevant than a global date alignment.
3. **Merging**: Whether you choose to use alignment or not, your wide-format table remains mergeable with other wide-format tables in DataSHIELD if they share the same `person_id` (or whichever merge column(s) you specify).

In short, `wideLongitudinal` provides a wide “spread” of repeated measurements within each row, and `completeTimePoints` ensures that these measurements line up on a shared calendar of dates. You can fine-tune these parameters to suit your analytic design.

# 2.2 Using the `get` method

The `get` method has several arguments that allow for the customization of the data that is retrieved. These arguments are:

- `table`: the name of the table to be retrieved.
- `symbol`: the name of the symbol that will be used to store the table in the server's R environment. If not specified, the table will be stored in a symbol with the same name as the table.
- `conceptFilter`: a numeric vector with the concept IDs to be retrieved. It can also be a single numeric value.
- `columnFilter`: a character vector with the names of the columns to be retrieved. It can also be a single character value.
- `personFilter`: the name of another symbol in the server's R environment which contains the person IDs to be retrieved by the current `get` operation. If not specified, all existing person IDs will be retrieved.
- `mergeColumn`: the name of the column that will be used to merge the table with another table. If not specified, the `person_id` column will be used as the default merge column.
- `dropNA`: a logical value indicating whether empty columns should be automatically filtered out from the table. The default value is `TRUE`.

Note that only the `table` parameter is mandatory for retrieving a table. However, it is highly recommended to utilize the available data filtering parameters, as they can significantly enhance the performance of operations and ensure that only the relevant data for your study is selected.

## 2.2.1 Applying basic filters

This part is quite straightforward. We use the `tables()` command to know which tables we can take, `columns()` to know what columns we have available, and `concepts()` to understand what concepts are present as entities in the table and select those we wish to work with.

Once we have identified the specific data we are interested in analyzing, we can apply the `columnFilter` and `conceptFilter` parameters to retrieve the desired `table`.

Let's proceed with an example.

We check the available tables:

```{r tables}
o$tables()
```

Let's say we want to retrieve some measurements from the `Measurement` table. We can use the `concepts()` command to see what concepts are available in this table:

```{r concepts, R.options = list(max.print = 10)}
o$concepts("measurement")
```

We have decided to retrieve the information related to the concept `3010421` (which corresponds to `pH of blood`). 

But we don't want to retrieve all the columns from the table, so we can use the `columns()` command to see what columns are available in the `Measurement` table:

```{r columns}
o$columns("measurement")
```

We are only interested in the `measurement_date` and `value_as_number` from the `Measurement` table related to the concept ID `3010421`. We can use the `get` method to retrieve this information:

```{r get_measurement_3000099}
o$get(table = "measurement",  # The table we want to retrieve
      symbol = "my_table",  # The name of the symbol that will store the table
      conceptFilter = 3010421,  # The concept we are interested in
      columnFilter = c("measurement_date", "value_as_concept_id"),  # The only columns we want to retrieve
      dropNA = FALSE  # Whether we want to drop empty columns or not
)

ds.summary("my_table")
```

Now this information is ready to be used within the DataSHIELD environment for any analysis or data manipulation operations that we want to perform. In the next chapter, we will see how to use the built-in functions of DataSHIELD's `dsBaseClient` package to combine it with other tables.

## 2.2.2 Selecting a merge column

When retrieving a table, it is possible to specify a column that will be used to merge the table with another table. This is done using the `mergeColumn` argument. For example, when retrieving the `Visit_detail` table, we can specify that the column `visit_occurrence_id` will be used to merge this table with the `Visit_occurrence` table, and it will perform the reshaping operation to make this merge possible.

For this example, we will retrieve the `Visit_detail` table related to the concept ID `4149943` (which corresponds to `Cardiac intensive care unit`, meaning that we are interested in retrieving only the visits to this unit):

```{r merge_column}
o$get(table = "visit_detail",
      conceptFilter = 4149943,  # `Cardiac intensive care unit`
      columnFilter = c("visit_detail_start_date", "visit_detail_end_date"),
      mergeColumn = "visit_occurrence_id")

ds.summary("visit_detail")
```

If no `mergeColumn` is specified, the `get` method will assume `person_id` as the default merge column, as this is the most common column that is used to merge tables in an OMOP CDM database.

## 2.2.3 Filtering by person IDs

The `personFilter` argument allows for the retrieval of data for a specific subset of person IDs. This can be useful when working with a large dataset and only needing to retrieve data for a specific group of individuals.

To use the `personFilter` parameter, we have to specify the symbol of another table that already exists in the R environment. This referenced table has to contain `person_id` values that may represent a subset of the `person_id` values in the table that we aim to retrieve. By applying this parameter, the retrieval process filters the target table to include only records corresponding to the `person_id` values present in the referenced table, thereby excluding any records associated with `person_id` values not contained in the referenced table.

Utilizing the subsetting functions provided by the `dsBaseClient` package can significantly enhance the versatility of filtering database records based on cohorts, conditions, or any other characteristics defining a subset of individuals. Furthermore, the `dsOMOP` package implements a disclosure control process that prevents the retrieval of tables containing information on a number of patients below a predefined threshold set in the server's configuration. This ensures the security and confidentiality of the data at all times.

Let's perform a simple example to illustrate the potential of this feature.

Suppose that, in this case, we are interested in analyzing data from patients who have records of the condition `Chest pain` (Concept ID `77670` in `Condition_occurrence`). We can save this information to a specific `symbol` and then pass it as the `personFilter` argument to the `get` method for the `Person` table:

```{r set_filter_criteria}
o$get(table = "condition_occurrence",
      conceptFilter = 77670,  # `Chest pain`
      columnFilter = "condition_start_date",
      symbol = "chest_pain_condition")

ds.summary("chest_pain_condition")
```

We stored the measurements under the `chest_pain_condition` symbol. Now we can use this table (referenced through the `chest_pain_condition` symbol) to filter the `Person` table and retrieve only the data related to these patients:

```{r filter_person_by_condition}
o$get(table = "person",
      symbol = "person_with_chest_pain",  # The name of the symbol that will store the table
      personFilter = "chest_pain_condition")  # The symbol that contains the only person IDs to be retrieved
```

Now we have a subset of the `Person` table that only contains the data related to the patients who have records of the condition `Chest pain`, stored in the `person_with_chest_pain` symbol. We can pass it to the `personFilter` argument of the `get` method for a third table to retrieve only the data related to these patients. For example, we can retrieve the `Observation` table using the `person_with_chest_pain` symbol as the `personFilter`, and it will only retrieve the data related to the the patients who are present in the `person_with_chest_pain` symbol:

```{r filter_observation_by_person, R.options = list(max.print = 10)}
o$get(table = "observation",
      symbol = "observation_from_person_with_chest_pain",
      columnFilter = "observation_date",
      personFilter = "person_with_chest_pain")

ds.summary("observation_from_person_with_chest_pain")
```

The `Observation` table we retrieved only contains all the observation records related to the patients who have records of the condition `Chest pain`.

This way, we can easily filter the data we are interested in and work with a subset of the data that is relevant to our study. This can also be done with cohorts, conditions, and any other characteristics that define a subset of individuals. It is done at query time, which makes data retrieval efficient.

In the next article, we will see how to use the `dsBaseClient` package to combine the data we have retrieved to create a combined dataset that can be used for analysis. 

```{r logout, echo = FALSE}
datashield.logout(conns)
```
