---
title: "Cohort Management"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Cohort Management}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(eval = FALSE)
```

## Overview

A cohort defines a set of persons of interest. dsOMOP v2 supports three ways to
obtain one:

1. **Pre-existing cohorts** stored in the results schema (e.g. generated by
   ATLAS or CohortGenerator).
2. **Spec-based cohorts** created on-the-fly from a structured specification.
3. **Workspace cohorts** built from an already-extracted `omop.table` with
   `ds.omop.cohort.from_table()`.

All three produce a server-side cohort table (carried client-side as a
`dsomop_cohort_handle`) that can scope exploration queries (the unified
`cohort =` argument) and filter extraction plans and recipes.

## Listing and Inspecting Cohorts

```{r list}
library(dsOMOPClient)

cohorts <- ds.omop.cohort.list(symbol = "omop", conns = conns)
# cohort_definition_id, cohort_definition_name, ...

defn <- ds.omop.cohort.definition(42, symbol = "omop", conns = conns)
# Full definition details for cohort 42
```

## Creating a Cohort from a Spec

`ds.omop.cohort.create()` takes a `spec` list that must contain a `type` and a
`concept_set`:

```{r create-condition}
diabetes <- ds.omop.cohort.create(
  spec = list(type = "condition", concept_set = c(201826)),
  cohort_id = 100,
  name = "diabetes_cohort",
  mode = "temporary",
  symbol = "omop",
  conns = conns
)
# Returns a dsomop_cohort_handle pointing at the server-side cohort table
```

### Supported Spec Types

| Type | Source Table | Notes |
|---|---|---|
| `condition` | condition_occurrence | Filter by condition concepts |
| `drug` | drug_exposure | Filter by drug concepts |
| `measurement` | measurement | Filter by concepts + optional value threshold |
| `observation` | observation | Filter by observation concepts |
| `procedure` | procedure_occurrence | Filter by procedure concepts |

### Measurement with Value Threshold

```{r create-measurement}
high_hba1c <- ds.omop.cohort.create(
  spec = list(
    type = "measurement",
    concept_set = c(3004410),
    value_threshold = list(op = ">=", value = 6.5)
  ),
  cohort_id = 101,
  name = "high_hba1c",
  mode = "temporary",
  symbol = "omop",
  conns = conns
)
```

### Persistent Cohorts

`mode = "persistent"` writes to the cohort schema so the cohort survives across
sessions, instead of the default session-scoped temporary table.

```{r create-persistent}
ds.omop.cohort.create(
  spec = list(type = "condition", concept_set = c(201826)),
  cohort_id = 200,
  name = "diabetes_persistent",
  mode = "persistent",
  symbol = "omop",
  conns = conns
)
```

## Combining Cohorts

`ds.omop.cohort.combine()` accepts the handles returned by `cohort.create()`
(or cohort table names) and applies a set operation. The combined result is
re-gated on its distinct-person count:

```{r combine}
both <- ds.omop.cohort.combine(
  op = "intersect",          # "union", "intersect", or "setdiff"
  cohort_a = diabetes,
  cohort_b = high_hba1c,
  new_name = "dm_and_high_hba1c",
  symbol = "omop",
  conns = conns
)
```

`"difference"` is accepted as an alias for `"setdiff"`. Every set operation is
gated on `nfilter.subset`; a result describing too few unique persons is blocked.

## Building a Cohort from a Workspace Table

Any server-side `omop.table` you have already produced -- the output of
`ds.omop.plan.execute()`, of a recipe, or of a manipulation verb
(`ds.omop.merge()` and friends) -- can be promoted to a reusable cohort. Only
the symbol **name** travels to the server; the server reads its distinct person
tokens, reverses them to original ids with the per-resource key, gates the count
(fail-closed), and materialises a size-checked cohort table. No identifier ever
leaves the server.

```{r from-table}
# Suppose a plan produced a features table under the symbol "F"
ds.omop.plan.execute(plan, out = c(features = "F"),
  symbol = "omop", conns = conns)

# Turn the persons in F into a cohort, then reuse it to scope exploration
coh <- ds.omop.cohort.from_table("F", symbol = "omop", conns = conns)
ds.omop.concept.prevalence("condition_occurrence", cohort = coh,
  symbol = "omop", conns = conns)
```

## The Unified `cohort =` Argument

Exploration aggregates accept a single `cohort =` argument that resolves any of:
a `dsomop_cohort_handle`, a numeric cohort definition id, or a server-side cohort
table name. This lets you run the same profiling within a subpopulation:

```{r cohort-arg}
ds.omop.column.stats("measurement", "value_as_number", concept_id = 3004410,
  cohort = diabetes, symbol = "omop", conns = conns)
ds.omop.missingness("measurement", cohort = diabetes,
  symbol = "omop", conns = conns)
ds.omop.crosstab("person", "gender_concept_id", "race_concept_id",
  cohort = diabetes, symbol = "omop", conns = conns)
ds.omop.value.counts("condition_occurrence", "condition_concept_id",
  cohort = 42, symbol = "omop", conns = conns)   # by definition id
```

## Using Cohorts in Plans

Reference a cohort in an extraction plan with `ds.omop.plan.cohort()`:

```{r plan-cohort}
plan <- ds.omop.plan()

# Use an existing cohort by definition id ...
plan <- ds.omop.plan.cohort(plan, cohort_definition_id = 42)

# ... or define one inline with a spec
plan <- ds.omop.plan.cohort(plan,
  spec = list(type = "condition", concept_set = c(201826)))

plan <- ds.omop.plan.baseline(plan,
  columns = c("gender_concept_id", "year_of_birth"))

ds.omop.plan.execute(plan, out = c(baseline = "D"),
  symbol = "omop", conns = conns)
```

`ds.omop.cohort.ref(id)` builds the same `cohort_definition_id` reference object
if you prefer to assemble the reference separately.

## Disclosure Controls

Every cohort operation -- creation, combination, and `from_table` promotion --
enforces `nfilter.subset` on the number of **unique persons** (counted as
`COUNT(DISTINCT person_id)`). If a cohort would contain fewer than the threshold,
the operation fails closed:

```
Error: Disclosive: unique persons (2) < nfilter.subset (3)
```

See the *Security* vignette for how an Opal administrator configures the
thresholds.
