---
title: "3. Integrating data into the workflow"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{integration}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r creating_object, include = FALSE}
library(DSI)
library(DSOpal)
library(dsBaseClient)
library(dsOMOPClient)

builder <- newDSLoginBuilder()
builder$append(
  server = "opal_demo",
  url = "https://opal-demo.obiba.org",
  user = "dsuser",
  password = "P@ssw0rd",
  profile = "omop"
)
logindata <- builder$build()
conns <- datashield.login(logins = logindata)

o <- ds.omop(
  connections = conns,
  resource = "omop_demo.mimiciv"
)
```

Having explored how to retrieve tables using the `get` method from `dsOMOPClient` and its customization parameters, it is now time to integrate these tables into the DataSHIELD workflow properly. To achieve this, we will primarily use two commands from the `dsBaseClient` package:

- `ds.merge`, to combine tables into a single data frame.
- `ds.dataFrameSubset`, to apply value filters to tables based on desired characteristics.

Make sure that you have the `dsBaseClient` package installed and loaded in your DataSHIELD environment:

```{r dsBaseClient, eval = FALSE}
library(dsBaseClient)
```

# 3.1 Merging tables

In this section, we will demonstrate how to combine the `Person` table and the `Measurement` table. This process involves retrieving the necessary data from both tables and then merging them based on a common identifier.

First, we retrieve the `Person` table:

```{r get_person}
o$get("person")

ds.summary("person")
```

Then, we retrieve the `Measurement` table. For this example, we will only retrieve the numerical value (`value_as_number`) and the date of the measurement (`measurement_date`) for the concept `Dyastolic blood pressure`, which has a concept ID of `3012888`:

```{r get_measurement}
o$get("measurement",
      columnFilter = c("value_as_number", "measurement_date"),
      conceptFilter = c(3012888))

ds.summary("measurement")
```

We can now merge the two tables based on the `person_id` column, using the conventional `ds.merge` command, which adheres to DataSHIELD's data disclosure rules:

```{r merging_tables}
ds.merge(
  x.name = "person",
  y.name = "measurement",
  by.x.names = "person_id",
  by.y.names = "person_id",
  all.x = TRUE,
  all.y = FALSE,
  sort = FALSE,
  suffixes = c(".x", ".y"),
  no.dups = TRUE,
  newobj = "merged",
  datasources = conns
)
```

We can see the `merged` table includes the columns from both the `Person` and `Measurement` tables:

```{r merged_summary}
ds.summary("merged")
```

This is how we can merge tables in DataSHIELD. We could also merge more tables by repeating the process with additional tables and the combined table.

## 3.1.1 Auxiliar functions and packages

Given the complexity that the previous operations could pose when having to work with multiple concepts and tables, it is recommended to use and develop auxiliary packages that encapsulate and automate the interaction with `dsOMOPClient`, which only acts as an interface with the database data, to meet the research needs of each specific case.

Below we can see an example of how `dsOMOPHelper`, a package that encapsulates the interaction with `dsOMOPClient`, can automatically join multiple tables in batch by the desired concepts using the `auto` function. 

For this example, we will retrieve the concepts `Tobacco use` (identified by the concept ID `4005823` in the `Observation` table) and `Chronic obstructive pulmonary disease` (identified by the concept ID `255573` in the `Condition_occurrence` table):

```{r auxiliar_packages} 
library(dsOMOPHelper)

h <- ds.omop.helper(connections = conns,
                    resource = "omop_demo.mimiciv",
                    symbol = "automatic_table")

h$auto(tables = c("observation", "condition_occurrence"),
       columns = c("observation_date", "condition_start_date"),
       concepts = c(4005823,  # `Tobacco use`
                    255573))  # `Chronic obstructive pulmonary disease`

ds.summary("automatic_table")
```

This is precisely why we encourage the community to develop their own packages on top of `dsOMOPClient`. Creating custom packages with specific workflows allows researchers and developers to tailor the functionality of `dsOMOPClient` to easily meet the specific needs of their projects. For more information about `dsOMOPHelper` and how it can automate the process of working with OMOP CDM databases in DataSHIELD, please visit [its GitHub repository](https://github.com/isglobal-brge/dsOMOPHelper).

# 3.2 Subsetting tables

Another fundamental operation in data selection and dataset construction that may be required for specific research contexts is the `ds.dataFrameSubset` operation from the `dsBaseClient` package. It is used to filter tables based on specific criteria, such as gender, age, or other characteristics. This operation can be used to determine a specific cohort of individuals based on their characteristics so that they can be used for further data retrieval and analysis.

Let's see with a simple example how we can filter, using the subset operation, a dataset to only include female individuals.

First, we get the `Person` table, which contains `gender_concept_id`, and verify that this column includes values for `female` and `male`.

```{r get_person_gender}
o$get("person")

ds.summary("person$gender_concept_id")
```

From this point on, we will work with the standard DataSHIELD procedures to filter the data. First, we create a `gender_filter` object that contains the value `female`:

```{r create_gender_filter}
ds.make(newobj = "gender_filter", toAssign = "c('female')")
```

Then, we use the `ds.dataFrameSubset` function to apply the `gender_filter` to the `gender_concept_id` column of the `Person` table:

```{r subset_by_gender}
ds.dataFrameSubset(
  df.name = "person",
  V1.name = "person$gender_concept_id",
  V2.name = "gender_filter",
  Boolean.operator = "==",
  newobj = "person_female",
  datasources = conns,
  notify.of.progress = FALSE
)
```

Thus, we can verify that the resulting table `person_female` only contains individuals with a `gender_concept_id` value of `female`:

```{r check_gender_subset}
ds.summary("person_female$gender_concept_id")
```

Once the `person_female` subset is obtained, we can pass it as the `personFilter` parameter to the `get` function to retrieve another table, in this case, `Measurement`. In this case, we are going to retrieve the measurements of `Mean blood pressure`, which has a concept ID of `3027598`. The resulting table will contain only the measurements from individuals with a `gender_concept_id` value of `female`:

```{r get_person_females}
o$get(table = "measurement",
      symbol = "measurement_female",
      columnFilter = c("value_as_number", "measurement_date"),
      conceptFilter = c(3027598),  # `Mean blood pressure`
      personFilter = "person_female")

ds.summary("measurement_female")
```

We can now merge the `person_female` and `measurement_female` tables based on the `person_id` column:

```{r merging_female_measurements}
ds.merge(
  x.name = "person_female",
  y.name = "measurement_female",
  by.x.names = "person_id",
  by.y.names = "person_id",
  all.x = TRUE,
  all.y = FALSE,
  sort = FALSE,
  suffixes = c(".x", ".y"),
  no.dups = TRUE,
  newobj = "merged_female",
  datasources = conns
)
```

We can see the merged table includes the columns from both the `Person` and `Measurement` tables, but only for individuals with a `gender_concept_id` value of `female`:

```{r check_female_measurements}
# General summary of the resulting table
ds.summary("merged_female")

# Summary of the `gender_concept_id` column to verify that it only includes `female` values
ds.summary("merged_female$gender_concept_id")
```

This is how we can use DataSHIELD's standard methods to subset tables obtained from OMOP CDM databases. We can achieve complex filters by applying more sophisticated `ds.dataFrameSubset` logics and propagating them to the desired tables through the `personFilter` parameter of the `get` method.

# Logout

We have arrived at the end of these vignettes. If we have finished working with our DataSHIELD session, it is a good practice to disconnect from the DataSHIELD server to avoid any potential issues: 

```{r logout}
datashield.logout(conns)
```
