---
title: "3. Workflow integration"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{integration}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r creating_object, include = FALSE}
library(DSI)
library(DSOpal)
library(dsBaseClient)
library(dsOMOPClient)

builder <- newDSLoginBuilder()
builder$append(
  server = "brge",
  url = "https://opal.isglobal.org/brge",
  user = "omop_demo",
  password = "P@ssw0rd"
)
logindata <- builder$build()
conns <- datashield.login(logins = logindata)

o <- ds.omop(
  connections = conns,
  resource = "omop_demo.mimiciv"
)
```

Having explored how to retrieve tables using the `get` method from `dsOMOPClient` and its customization parameters, it is now time to integrate these tables into the DataSHIELD workflow. To achieve this, we will primarily use two commands from the `dsBaseClient` package:

- `ds.merge`, to combine tables into a single data frame.
- `ds.dataFrameSubset`, to apply value filters to tables based on desired characteristics.

# 3.1 Merging tables

In this section, we will demonstrate how to combine the `Person` table and the `Measurement` table. This process involves retrieving the necessary data from both tables and then merging them based on a common identifier.

First, we retrieve the `Person` table:

```{r get_person}
o$get("person")
ds.summary("person")
```

Then, we retrieve the `Measurement` table:

```{r get_measurement}
o$get("measurement",
      columnFilter = c("value_as_number", "measurement_date"),
      conceptFilter = c(3000099))

ds.summary("measurement")
```

We can now merge the two tables based on the `person_id` column, using the conventional `ds.merge` command, which adheres to DataSHIELD's data disclosure rules:

```{r merging_tables}
ds.merge(
  x.name = "person",
  y.name = "measurement",
  by.x.names = "person_id",
  by.y.names = "person_id",
  all.x = TRUE,
  all.y = FALSE,
  sort = FALSE,
  suffixes = c(".x", ".y"),
  no.dups = TRUE,
  newobj = "person_measurement",
  datasources = conns
)
```

We can see the merged table includes the columns from both the `Person` and `Measurement` tables:

```{r merged_summary}
ds.summary("person_measurement")
```

This is how we can merge tables in DataSHIELD. We could also merge more tables by repeating the process with additional tables and the combined table.

## 3.1.1 Auxiliar functions and packages

Given the complexity of the previous example, it is recommended to use and develop auxiliary packages that encapsulate and automate the interaction with `dsOMOPClient`, which only acts as an interface with the database data, to meet the research needs of each specific case.

Below we can see an example of how the `dsOMOPHelper` package can automatically join multiple tables in batch by the desired concepts using the `auto` function:

```{r auxiliar_packages} 
library(dsOMOPHelper)

h <- ds.omop.helper(connections = conns,
                    resource = "omop_demo.mimiciv",
                    symbol = "resulting_table")

h$auto(tables = c("measurement", "condition_occurrence"),
       columns = c("date", "start_date"),
       concepts = c(3000099, 77670))

ds.summary("resulting_table")
```

This is precisely why we encourage the community to develop their own packages on top of `dsOMOPClient`. By creating custom packages, researchers and developers can tailor the functionality to meet the specific needs of their projects, enhancing the efficiency and effectiveness of their analyses.

For more information about `dsOMOPHelper` and how it can streamline the process of working with OMOP CDM databases in DataSHIELD, please visit [its GitHub repository](https://github.com/isglobal-brge/dsOMOPHelper).

# 3.2 Subsetting tables

Another fundamental operation in data selection and dataset construction that fits specific research needs in DataSHIELD is the `ds.dataFrameSubset` operation from the `dsBaseClient` package.

Let's see with a simple example how we can filter, using the subset operation, a dataset to only include female individuals.

First, we take the `Person` table, which contains `gender_concept_id`, and verify that this column includes values for `female` and `male`.

```{r get_person_gender}
o$get("person")
ds.summary("person$gender_concept_id")
```

From this point on, we will work with the standard DataSHIELD procedures to filter the data. First, we create a `gender_filter` object that contains the value `female`:

```{r create_gender_filter}
ds.make(newobj = "gender_filter", toAssign = "c('female')")
```

Then, we use the `ds.dataFrameSubset` function to apply the `gender_filter` to the `gender_concept_id` column of the `Person` table:

```{r subset_by_gender}
ds.dataFrameSubset(
  df.name = "person",
  V1.name = "person$gender_concept_id",
  V2.name = "gender_filter",
  Boolean.operator = "==",
  newobj = "person_female",
  datasources = conns,
  notify.of.progress = FALSE
)
```

Thus, we can verify that the resulting table `person_female` only contains individuals with a `gender_concept_id` value of `female`:

```{r check_gender_subset}
ds.summary("person_female$gender_concept_id")
```

Once the `person_female` subset is obtained, we can pass it as the `personFilter` parameter to the `get` function to retrieve another table, in this case, `Measurement`. The resulting table will contain only the measurements from individuals with a `gender_concept_id` value of `female`:

```{r get_person_females}
o$get(table = "measurement",
      symbol = "measurement_female",
      columnFilter = c("value_as_number", "measurement_date"),
      conceptFilter = c(3000099),
      personFilter = "person_female")

ds.summary("measurement_female")
```

We can now merge the `person_female` and `measurement_female` tables based on the `person_id` column:

```{r merging_female_measurements}
ds.merge(
  x.name = "person_female",
  y.name = "measurement_female",
  by.x.names = "person_id",
  by.y.names = "person_id",
  all.x = TRUE,
  all.y = FALSE,
  sort = FALSE,
  suffixes = c(".x", ".y"),
  no.dups = TRUE,
  newobj = "merged_female",
  datasources = conns
)
```

We can see the merged table includes the columns from both the `Person` and `Measurement` tables, but only for individuals with a `gender_concept_id` value of `female`:

```{r check_female_measurements}
ds.summary("merged_female")
ds.summary("merged_female$gender_concept_id")
```

This is how we can use DataSHIELD's standard methods to subset tables obtained from OMOP CDM databases. We can achieve complex filters by applying more sophisticated `ds.dataFrameSubset` logics and propagating them to the desired tables through the `personFilter` parameter of the `get` method.
