---
title: "Data Extraction with the Plan DSL"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Data Extraction with the Plan DSL}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(eval = FALSE)
```

## Overview

The **extraction plan DSL** specifies exactly what data to pull from the OMOP
CDM. A plan is a serializable object describing a cohort, one or more outputs,
their representations, and global options. Plans are built client-side, validated
against the server schema, and executed to produce server-side data frames
(`omop.table` objects) that you analyse with `dsBaseClient`.

For higher-level, named, save/loadable extractions, the **recipe** API wraps the
plan (covered at the end of this vignette).

## Building a Plan

```{r basic}
library(dsOMOPClient)

plan <- ds.omop.plan()
# An empty plan. Concept translation and federation-wide concept factors are
# ON by default (translate_concepts = TRUE, factor_concepts = TRUE).
```

### Baseline demographics

`ds.omop.plan.baseline()` adds a person-level output of demographic columns from
the `person` table, plus optional derived columns (e.g. `age_at_index`):

```{r baseline}
plan <- ds.omop.plan.baseline(plan,
  columns = c("gender_concept_id", "year_of_birth", "race_concept_id"),
  derived = c("age_at_index"),
  name = "demographics")
```

To merge columns from several person-linked tables into one person-level frame,
use `ds.omop.plan.person_level()`:

```{r person-level}
plan <- ds.omop.plan.person_level(plan,
  tables = list(
    person = c("gender_concept_id", "year_of_birth"),
    observation_period = c("observation_period_start_date",
                           "observation_period_end_date")),
  name = "person_data")
```

### Event-level data

`ds.omop.plan.events()` extracts rows from a clinical event table:

```{r events}
plan <- ds.omop.plan.events(plan,
  name = "conditions",
  table = "condition_occurrence",
  columns = c("person_id", "condition_concept_id", "condition_start_date"),
  concept_set = c(201826, 255573, 4329847),
  representation = list(format = "long"))
```

### Outcomes and feature sets

`ds.omop.plan.outcome()` is a convenience wrapper for a binary condition outcome
(it adds an event output with the `features` representation):

```{r outcome}
plan <- ds.omop.plan.outcome(plan, name = "copd", concept_set = c(255573))
```

`ds.omop.plan.features()` builds a person-level feature set from explicit feature
specs (see the `omop.feature.*` constructors):

```{r features}
plan <- ds.omop.plan.features(plan,
  name = "hba1c_features",
  table = "measurement",
  specs = list(
    omop.feature.mean_value(concept_set = c(3004410)),
    omop.feature.latest_value(concept_set = c(3004410)),
    omop.feature.sd_value(concept_set = c(3004410))))
```

## Representations

Each event output uses one of three representations:

- **`long`** (default): one row per event, full temporal detail.
- **`wide`**: pivots on the concept column; each concept becomes a column group.
- **`features`**: aggregates to one row per person (counts, ever-flags, and for
  numeric values mean/min/max/sd, etc.).

```{r repr}
plan <- ds.omop.plan.events(plan, name = "labs_wide",
  table = "measurement", concept_set = c(3004410),
  representation = list(format = "wide"))
```

## Maximum-Flexibility Extraction

### Row-level filter DSL

Beyond a concept set, `ds.omop.plan.events()` accepts a `filters` argument that
takes a **nested AND/OR tree of row-level predicates**. Build leaves with the
`omop_filter*` constructors and combine them with
`omop_filter_group(..., operator = "AND" | "OR")`:

```{r row-filters}
filt <- omop_filter_group(
  omop_filter(type = "custom", level = "row",
    params = list(var = "value_as_number", op = "between", value = c(6.5, 12))),
  omop_filter_group(
    omop_filter_date_range(start = "2020-01-01", end = "2023-12-31"),
    operator = "OR"),
  operator = "AND")

plan <- ds.omop.plan.events(plan, name = "labs_filtered",
  table = "measurement", concept_set = c(3004410),
  filters = filt)
```

Disclosure note: the custom filter DSL is fail-closed. Range and membership
operators (`>=`, `>`, `<=`, `<`, `between`, `in`, `not_in`, `is_null`,
`not_null`) and the pre-validated `value_bin` family are allowed because they can
only narrow a population. Exact-match predicates (`==`, `!=`) on an arbitrary
column are **rejected server-side** -- a single arbitrary-column equality is the
canonical fingerprinting primitive -- and filtering on the person token or any
protected/identifier column is also rejected. The per-patient gate still runs on
the filtered result, so a filter can never bypass suppression. For value
thresholds snapped to disclosure-safe bin edges, prefer
`ds.omop.safe.cutpoints()` / `ds.omop.safe.filter.value()`, which return a
sanctioned `value_bin` filter.

### Unit / type concept-column scoping

By default a table is profiled and filtered on its domain concept column. Pass
`concept_col` to scope to a different concept column instead -- e.g. a unit
(`unit_concept_id`), a record-type (`*_type_concept_id`), or a value concept
(`value_as_concept_id`):

```{r concept-col}
plan <- ds.omop.plan.events(plan, name = "labs_by_unit",
  table = "measurement", concept_set = c(8840),  # mg/dL
  concept_col = "unit_concept_id")
```

### Visit linkage

`visit_filter` restricts events to those occurring during a visit of a given
type, via an `EXISTS` join on `visit_occurrence_id`:

```{r visit}
plan <- ds.omop.plan.events(plan, name = "inpatient_dx",
  table = "condition_occurrence", concept_set = c(201826),
  visit_filter = list(visit_concept_id = c(9201)))  # inpatient
```

## Concept Translation

`translate_concepts` is **ON by default**. Concept-ID columns are replaced with
human-readable concept names from the vocabulary, so `condition_concept_id =
201826` is returned as `"Type 2 diabetes mellitus"`. Toggle it (and the related
options) with `ds.omop.plan.options()`:

```{r options}
plan <- ds.omop.plan.options(plan,
  translate_concepts = TRUE,
  factor_concepts = TRUE,
  block_sensitive = TRUE)
```

`factor_concepts = TRUE` (the default) makes execution recode concept columns as
federation-wide factors with a shared level ordering, which is what lets pooled
`ds.glm`, `ds.glmSLMA`, and `ds.table` behave correctly across servers.

## Plan Lifecycle

### Preview (server-side dry run)

`ds.omop.plan.preview()` sends the plan for a dry run that returns, **per
server**, each output's expected columns, an estimated person count, and a
disclosure flag -- *without creating any data*:

```{r preview}
preview <- ds.omop.plan.preview(plan, symbol = "omop", conns = conns)

# Per-server, per-output details:
preview$server1$outputs$conditions$columns
preview$server1$outputs$conditions$n_persons    # banded, not exact
preview$server1$outputs$conditions$disclosive   # TRUE if below nfilter.subset
```

The estimated `n_persons` is **banded** (rounded down to a multiple of
`band_width`) so the preview never reveals an exact supra-threshold count that
could be used for differencing; an output whose person count is below
`nfilter.subset` is flagged `disclosive = TRUE`. `ds.omop.plan.validate()` calls
the same server endpoint; use it for a pass/fail check and `preview` to inspect
the expected structure.

### Execute

`ds.omop.plan.execute()` runs the plan and assigns each output into the
DataSHIELD session under the symbols given in `out`. `out` is a **named**
character vector mapping each plan output name to a server-side symbol:

```{r execute}
ds.omop.plan.execute(plan,
  out = c(demographics = "D", conditions = "E"),
  symbol = "omop", conns = conns)

# D and E are now server-side omop.table data frames
dsBaseClient::ds.summary("D")
dsBaseClient::ds.dim("E")
```

For very large extractions, `output_mode = "staged"` writes outputs to Parquet
on the server and assigns `FlowerDatasetDescriptor` objects instead of
in-memory data frames (requires `arrow` server-side).

## The Recipe API (named, reusable extractions)

A **recipe** is a higher-level, named container of populations, variables, and
outputs that compiles down to a plan. It is the most ergonomic path and the one
that supports saving/loading.

```{r recipe-build}
rec <- omop_recipe()
rec <- recipe_add_variable(rec,
  omop_variable(table = "person", column = "gender_concept_id",
                format = "sex_mf"))
rec <- recipe_add_variable(rec,
  omop_variable(table = "measurement", concept_id = 3004410,
                format = "mean",
                # per-variable row filters and a per-variable time window
                filters = list(omop_filter_date_range(start = "2020-01-01",
                                                      end = "2023-12-31")),
                time_window = list(start = -365, end = 0),
                concept_col = "value_as_concept_id"))
rec <- recipe_add_output(rec,
  omop_output(name = "study", type = "features"))
```

Per-variable `filters` are forwarded as row-level predicates and `time_window`
(in days, relative to the cohort index date) becomes the variable's temporal
window when the recipe is compiled to a plan.

### Validate, preview, execute

```{r recipe-run}
recipe_validate(rec, symbol = "omop", conns = conns)
recipe_preview(rec, symbol = "omop", conns = conns)

# Ergonomic execute: with no `out`, symbols are derived from the output names
# (here "study" -> "D_study"). This is the simplest single-output form.
recipe_execute(rec, symbol = "omop", conns = conns)

# Or assign explicit symbols, exactly like ds.omop.plan.execute():
recipe_execute(rec, out = c(study = "D"), symbol = "omop", conns = conns)
```

### Save and load (YAML / JSON)

Recipes round-trip to disk; the format is inferred from the file extension (or
set `format =` explicitly):

```{r recipe-io}
recipe_save(rec, "study.yaml")        # or "study.json"
rec2 <- recipe_load("study.yaml")

# Or work with strings instead of files:
yaml_txt <- recipe_export_yaml(rec)
rec3 <- recipe_import_yaml(yaml_txt)
```

You can also emit runnable R for the whole recipe with `recipe_to_code(rec)`.
