---
title: "Multi-Server Federated Analysis"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Multi-Server Federated Analysis}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(eval = FALSE)
```

## Overview

dsOMOP v2 is designed for federated analysis across multiple DataSHIELD servers,
each hosting an OMOP CDM database. Servers may differ in CDM version, available
tables and columns, vocabulary version, and data volume. dsOMOP handles these
differences through runtime schema introspection and plan harmonization.

## Connecting to Multiple Servers

```{r connect}
library(dsOMOPClient)
library(DSI)
library(DSOpal)

builder <- DSI::newDSLoginBuilder()
builder$append(server = "hospital_a", url = "https://opal-a.example.org",
  user = "analyst", password = "secret_a", resource = "project.omop_cdm")
builder$append(server = "hospital_b", url = "https://opal-b.example.org",
  user = "analyst", password = "secret_b", resource = "project.omop_cdm")
builder$append(server = "hospital_c", url = "https://opal-c.example.org",
  user = "analyst", password = "secret_c", resource = "project.omop_cdm")

login_data <- builder$build()
conns <- DSI::datashield.login(logins = login_data)
```

## Attaching Resources

`ds.omop.connect()` accepts a single resource name (applied to every server) or
a named list mapping servers to resources:

```{r attach}
# Same resource on all servers
ds.omop.connect(resource = "project.omop_cdm", conns = conns, symbol = "omop")

# Or a per-server mapping
ds.omop.connect(
  resource = list(
    hospital_a = "project.omop_cdm",
    hospital_b = "project.omop_cdm",
    hospital_c = "project.omop_cdm"),
  conns = conns, symbol = "omop")
```

## Schema Comparison

Before building a plan, compare schemas across servers with `ds.omop.compare()`:

```{r compare}
comparison <- ds.omop.compare(symbol = "omop", conns = conns)

comparison$common_tables   # tables available on ALL servers
comparison$server_only     # tables unique to particular servers
comparison$column_diffs    # which columns are missing where
```

## Plan Harmonization

When schemas differ, harmonize the plan before execution with
`ds.omop.plan.harmonize()`:

### Intersection mode (default)

Keeps only tables and columns present on every server:

```{r harmonize-intersect}
plan <- ds.omop.plan()
plan <- ds.omop.plan.person_level(plan,
  tables = list(
    person = c("gender_concept_id", "year_of_birth", "race_concept_id"),
    observation_period = c("observation_period_start_date")))

# If a server lacks race_concept_id, harmonize drops it everywhere
plan <- ds.omop.plan.harmonize(plan, mode = "intersection",
  symbol = "omop", conns = conns)
```

### Union-with-missing mode

Keeps all columns, filling those absent on a server with `NA`:

```{r harmonize-union}
plan <- ds.omop.plan.harmonize(plan, mode = "union_with_missing",
  symbol = "omop", conns = conns)
```

## Status Checking

```{r status}
status <- ds.omop.status(symbol = "omop")
# Per-server connection/handle status
```

## Federation-Wide Concept Factors

When a plan executes in memory mode with `factor_concepts = TRUE` (the default),
dsOMOP runs a cross-server coordination step after extraction: it collects each
server's disclosure-safe `_concept_id` levels, computes their union in one
deterministic order client-side, and broadcasts that ordering back so every
server recodes the columns as factors with identical level coding. A value
present on only some sites becomes an empty level on the others. This shared
coding is what makes pooled `ds.glm`, `ds.glmSLMA`, and `ds.table` behave
correctly on the federated factor.

## Federated Analysis Workflow

```{r workflow}
# 1. Connect and attach
conns <- DSI::datashield.login(logins = login_data)
ds.omop.connect(resource = "project.omop_cdm", conns = conns, symbol = "omop")

# 2. Compare schemas
comparison <- ds.omop.compare(symbol = "omop", conns = conns)

# 3. Build a plan on common ground
plan <- ds.omop.plan()
plan <- ds.omop.plan.cohort(plan,
  spec = list(type = "condition", concept_set = c(201826)))
plan <- ds.omop.plan.baseline(plan,
  columns = c("gender_concept_id", "year_of_birth"))
plan <- ds.omop.plan.features(plan, name = "labs", table = "measurement",
  specs = list(omop.feature.mean_value(concept_set = c(3004410))))

# 4. Harmonize, preview, execute
plan <- ds.omop.plan.harmonize(plan, mode = "intersection",
  symbol = "omop", conns = conns)
ds.omop.plan.preview(plan, symbol = "omop", conns = conns)
ds.omop.plan.execute(plan, out = c(baseline = "D", labs = "L"),
  symbol = "omop", conns = conns)

# 5. Federated modelling with dsBaseClient (concept columns are shared factors)
dsBaseClient::ds.glm(
  formula = "outcome ~ gender_concept_id + year_of_birth",
  data = "D", family = "binomial", datasources = conns)

# 6. Clean up
ds.omop.disconnect(symbol = "omop", conns = conns)
DSI::datashield.logout(conns)
```

## Version-Agnostic Design

dsOMOP v2 does not hard-code OMOP CDM version tables. It discovers available
tables and columns at runtime, so:

- CDM 5.3 and 5.4 databases work without configuration changes.
- Custom extensions to the CDM are discovered automatically.
- Missing optional tables are handled gracefully rather than as errors.

Each server's capabilities are introspected on connect and cached client-side to
avoid repeated round-trips.
