---
title: "Security and Disclosure Controls"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Security and Disclosure Controls}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(eval = FALSE)
```

## Overview

dsOMOP v2 enforces DataSHIELD disclosure controls at every level.
These controls are self-contained — dsOMOP does **not** depend on dsBase
for disclosure checking, which avoids version coupling.

All disclosure settings are read from DataSHIELD server options,
configured by the server administrator.

## Disclosure Settings

### nfilter.subset (Minimum Persons)

**The most critical control.** Any query result or cohort must contain
at least `nfilter.subset` unique persons.

- Default: 3
- Enforced using `COUNT(DISTINCT person_id)`
- Blocks: table extraction, cohort creation, cohort combination

```
# Server-side option
options(nfilter.subset = 5)
```

If a query would return data for fewer persons:

```
Error: Disclosive: unique persons (2) < nfilter.subset (3)
```

### nfilter.tab (Minimum Cell Count)

Minimum count for frequency distributions. Cells below this threshold
are replaced with `NA`.

- Default: 3
- Affects: `ds.omop.valueCounts()`, `catalogTableStats()`

### nfilter.levels.max (Maximum Factor Levels)

Maximum number of distinct values allowed for a categorical column.

- Default: 40
- Prevents extraction of high-cardinality identifiers

### nfilter.levels.density

Maximum ratio of levels to observations. Prevents near-unique columns
from being returned.

- Default: 0.33
- Formula: `n_levels / n_rows <= nfilter.levels.density`

### datashield.privacyControlLevel

Controls which operations are permitted:

| Level | Allowed Operations |
|---|---|
| `permissive` | All operations, including persistent cohort writes |
| `banana` | Most operations, including set difference |
| `avocado` | Standard operations |
| `carrot` | Restricted |
| `non-permissive` | Most restrictive |

Default: `"non-permissive"`

## Where Controls Are Enforced

### Table Extraction (.extractTable)

Before returning any data:

1. Counts unique persons in the query result
2. Blocks if below `nfilter.subset`
3. Validates factor levels against `nfilter.levels`

### Cohort Creation (.cohortCreate)

Before materializing a cohort:

1. Counts unique `subject_id` values
2. Blocks if below `nfilter.subset`

### Cohort Combination (.cohortCombine)

1. Counts unique persons in combined result
2. Set difference operations require `banana` privacy level

### Catalog Functions

- `tableStats`: Suppresses counts below `nfilter.subset`
- `valueCounts`: Suppresses cells below `nfilter.tab`, checks levels
- `searchConcepts`: Read-only vocabulary search (no patient data)

### Plan Preview (.planPreview)

Returns aggregate metadata. Each output is flagged:

```{r preview-flags}
preview <- ds.omop.plan.preview(plan, symbol = "omop", conns = conns)
# preview$outputs$conditions$disclosive == TRUE
# if the person count is below nfilter.subset
```

### Plan Execution (.planExecute)

Each output is individually checked. Disclosive outputs are skipped
with a warning rather than failing the entire plan.

## Input Validation

dsOMOP validates all user-provided strings to prevent SQL injection:

### Table and Column Names

- Must match `^[a-zA-Z_][a-zA-Z0-9_]*$`
- Maximum 128 characters
- Invalid names are rejected before SQL construction

### String Values

- Maximum 1000 characters by default
- No raw SQL is accepted from the client

### Concept IDs

- Cast to `as.integer()` before use in SQL
- Invalid values cause type coercion errors

## SQL Safety

dsOMOP never constructs SQL from raw user strings. All SQL is built
using parameterized patterns:

- Table/column names go through `adapter$quote_ident()`
- Literal values go through `adapter$quote_literal()`
- Concept IDs are cast to integer
- Filter DSL is compiled recursively with operator whitelisting

## Server Administrator Guide

### Recommended Settings

```r
# In Opal DataSHIELD options:
options(
  nfilter.subset = 5,           # Minimum unique persons
  nfilter.tab = 5,              # Minimum cell count
  nfilter.levels.max = 40,      # Max factor levels
  nfilter.levels.density = 0.33, # Max level density
  datashield.privacyControlLevel = "banana"
)
```

### DATASHIELD File

The `dsOMOP/DATASHIELD` file declares all methods and their default
option values. Opal reads this file during package installation:

```
Options:
  default.nfilter.subset=3
  default.nfilter.tab=3
  default.nfilter.levels.max=40
  default.nfilter.levels.density=0.33
  default.datashield.privacyControlLevel=non-permissive
```

Administrators can override these defaults in the Opal admin panel.

### Privacy Level Hierarchy

Operations are gated by privacy level. The hierarchy (most to least
permissive):

```
permissive > banana > avocado > carrot > non-permissive
```

Higher-risk operations require more permissive settings:

- **permissive**: Persistent cohort writes
- **banana**: Set difference (complement) operations
- **avocado**: Standard extraction and cohort creation
- **non-permissive**: Read-only catalog operations
