---
title: "The OMOP Studio Recipe Builder"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{The OMOP Studio Recipe Builder}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(eval = FALSE)
```

## What is OMOP Studio?

**OMOP Studio** is a point-and-click Shiny app for building and running
federated data extractions against OMOP CDM databases — no R scripting
required. You assemble a **recipe** by clicking (which cohort, which
variables, which output shape), and the Studio compiles it to a **plan** and
executes it on each DataSHIELD server.

The golden rule never changes: **row-level data never leaves the server.**
The recipe materialises result tables *on the nodes*; the Studio only ever
shows you disclosure-safe metadata (dimensions, column names). You then run
ordinary DataSHIELD analysis functions (`ds.glm`, `ds.table`, `ds.summary`, …)
on those server-side tables.

```
   You build a RECIPE  ──►  compiled to a PLAN  ──►  executed on each SERVER
   (cohort + variables          (recipe_to_plan)         (data stays there;
    + outputs, in the UI)                                  you get back only
                                                           a symbol + metadata)
```

## Launching the Studio

First establish a DataSHIELD session and an OMOP handle (the session symbol is
`"omop"` by default), then call `ds.omop.studio()`:

```{r launch}
library(dsOMOPClient)
library(DSI)

# 1. Log in to your DataSHIELD server(s)
builder <- newDSLoginBuilder()
builder$append(server = "server1", url = "https://opal.example.org",
               resource = "project.omop_cdm", driver = "OpalDriver")
conns <- datashield.login(builder$build())

# 2. Attach the OMOP handle — this creates the "omop" session the Studio uses
ds.omop.connect(resource = "project.omop_cdm", conns = conns)

# 3. Launch the Studio against that session
ds.omop.studio(symbol = "omop")
```

The app opens in your browser. Everything below happens in the **Builder**
tab, **Recipe** sub-tab.

## The recipe mental model

A recipe answers three questions:

| Question | In the Studio | Recipe piece |
|----------|---------------|--------------|
| **Who?** (the cohort) | *Add Filter* / *Group Filters* | population filters |
| **What?** (the columns) | *Add Variable* / *Block* / *Derived* | variables |
| **What shape?** (the table) | *Add Output* | outputs |
| **How?** (global behaviour) | *Plan Options* | options |

You add these in any order; the **Recipe Contents** card on the right shows
everything you have so far, and **Generated Code** mirrors it as reproducible
R you can copy.

## Building the cohort — *Add Filter*

Open **Add Filter**, pick a *Filter Type*, fill in its fields, click
**Add Filter**. The builder exposes all of the disclosure-safe cohort filters:

| Filter | Restricts the cohort to people who… |
|--------|--------------------------------------|
| **Sex** | are male / female |
| **Age Range** | are within an inclusive age band |
| **Age Group** | fall in one of several age bands (OR) |
| **Has Concept** | have ≥ N records of a concept in a table |
| **Concept Count** | have ≥ N records of a concept (occurrence threshold) |
| **Not Has Concept** | have **no** records of a concept |
| **Has Measurement** | have a measurement, optionally within a value range |
| **Missing Measurement** | have **no** non-null value for a measurement |
| **Prior Observation** | have ≥ N days of observation *before* index |
| **Follow-up** | have ≥ N days of observation *after* index |
| **Visit Count** | have ≥ N visits (optionally of a visit type) |
| **Date Range** | (row-level) keep only events in a calendar window |

Concept-based filters take a plain numeric **Concept ID** (the same value you
see in the Explore tab). Filters with a date or age band are constrained
server-side to safe minimum widths (age ≥ 5 years, dates ≥ 30 days).

### Combining filters with AND / OR — *Group Filters*

By default, every filter you add is combined with **AND**. To express richer
logic, open **Group Filters**:

1. Tick two or more filters from the list.
2. Choose **AND** (all must match) or **OR** (any may match).
3. Optionally name the group, then click **Combine into Group**.

The selected filters are wrapped into a single group in the Recipe Contents
list (shown as `[AND group]` / `[OR group]`). Groups compile to a nested
`AND`/`OR` filter tree on the cohort.

## Building the columns — *Add Variable*

Open **Add Variable**:

1. **Search a concept** in the picker (type a name or paste an ID) and click
   the match.
2. Choose the **Table** it comes from.
3. Choose a **Format** — what to compute per person (see the table below).
4. Optionally tick **Include descendant concepts** to match the concept *and*
   all of its hierarchical children (resolved server-side).
5. Click **Add Variable**.

### Variable formats

| Format | Produces (per person) |
|--------|------------------------|
| **Raw** | the raw event rows (use with a *long* output) |
| **Binary (0/1)** | 1 if the person has any matching record, else 0 |
| **Count** | number of matching records |
| **First / Last Value** | value from the earliest / most recent record |
| **Mean / Min / Max** | summary of a numeric value column |
| **Sum** | sum of a numeric column (e.g. days_supply) |
| **Std Dev / Coeff. of Variation** | spread of a numeric value |
| **Slope (trend)** | linear trend of a value over time |
| **Distinct Concepts** | number of distinct concepts in the table |
| **Drug Duration** | exposure length in days |
| **Duration Sum (days)** | total of end − start across records |
| **Abnormal High / Low (count)** | records outside the reference range |
| **Max / Mean Gap (days)** | gap between consecutive events |
| **Time Since** | days/months from the last event to a reference |
| **Binned** | *currently behaves as Binary at extraction time* |

> **Tip:** value-based formats (Mean, Std Dev, Slope, Sum…) default to a value
> column appropriate to the table (`value_as_number`, `days_supply`, …). If you
> pick one on a table that has no such column the result will be empty — use
> the **edit** pencil on the variable to set a different *Value Source*.

### Many concepts at once — *Add Variable Block*

Open **Add Variable Block** to paste a comma-separated list of concept IDs and
apply one format to all of them in a single step. It also supports
**Include descendant concepts**.

### Person-level / score variables — *Add Derived Variable*

Some variables need no concept at all — they are computed from the person and
observation tables. Open **Add Derived Variable**, pick a type, name it, and
click **Add Derived Variable**:

| Derived variable | Meaning |
|------------------|---------|
| **Age** | age at a reference date |
| **Sex (M/F)** | sex label from gender_concept_id |
| **Observation duration / Prior obs / Follow-up** | observation-window lengths in days |
| **Demographic missingness** | count of missing demographic fields |
| **Charlson / CHADS₂ / CHA₂DS₂-VASc** | standard comorbidity / stroke-risk scores |
| **DCSI / HFRS** | diabetes-complications / hospital-frailty scores |

## Choosing the shape — *Add Output*

Open **Add Output**, give it a name, and pick an **Output Type**:

| Output | Shape |
|--------|-------|
| **Wide** | one row per person, variables as columns (the usual choice) |
| **Baseline** | demographics + derived fields, one row per person |
| **Long** | raw event rows (many rows per person) |
| **Joined Long** | long events joined across tables |
| **Features / Covariates Sparse** | sparse feature matrix |
| **Survival** | event indicator + time-to-event (no calendar dates) |
| **Intervals** | start/end day intervals relative to index |

For **Long / Joined Long** outputs you also get event-level knobs:

- **Date handling** — `relative` (day offsets from index, the safe default),
  `binned` (week/month/year), or `remove`. Raw *absolute* dates are **not**
  offered here because they require explicit server permission.
- **Event selection** — keep only the **first** or **last N** qualifying
  events per person.

A **Staged (Parquet)** checkbox writes results to server-side Parquet files
(via `FlowerDatasetDescriptor`) instead of in-memory data frames, for very
large extractions.

## Global behaviour — *Plan Options*

Open **Plan Options** to set extraction-wide options:

| Option | Effect |
|--------|--------|
| **Translate concept IDs to labels** | Replaces numeric `*_concept_id` columns with human-readable concept names. **Turn this on for categorical analysis.** Off by default. |
| **Block sensitive columns** | Excludes exact dates / free-text columns (on by default; recommended). |
| **Harmonize concept factors across servers** | Recodes `*_concept_id` columns as factors with identical level coding on every node, so pooled `ds.table` / `ds.glm` align (on by default). |
| **Minimum persons per cell** | Extra small-count suppression threshold (blank = server default). |

> **Categorical data made readable.** With *Translate concept IDs to labels*
> ON, a `condition_concept_id` of `201820` comes back as
> `"diabetes_mellitus"` instead of an opaque integer — and, combined with
> factor harmonisation, you can feed it straight into `ds.table()` /
> `ds.glm()` for pooled categorical analysis.

## Review, generate, run

The right-hand panel updates live:

- **Recipe Contents** — every variable, filter, group, and output. Use the
  pencil to edit any item or the ✕ to remove it.
- **Schema Preview** — the columns your wide output will produce.
- **Generated Code** — the equivalent `dsOMOPClient` R script. Copy it with the
  copy button to reproduce the extraction from a console, or **Export** the
  recipe as JSON/YAML.
- **Run** (the ▶ button) — executes the recipe on the connected servers.

After **Run**, the **Execution Results** card shows, per server: whether each
result table was **assigned**, its **dimensions** (row × column counts), and
its **column names**. It never shows individual rows — that is the
non-disclosive guarantee in action. From here, analyse the server-side tables
with standard DataSHIELD functions.

## A worked example, end to end

**Goal:** a wide table of female patients with Type 2 diabetes, with their
age, Charlson score, a hypertension flag, and mean HbA1c — with readable
labels.

1. **Cohort.** *Add Filter* → **Sex** = Female. *Add Filter* → **Has Concept**
   = `201820` (Type 2 diabetes) in `condition_occurrence`.
2. **Derived variables.** *Add Derived Variable* → **Age**. Again → **Charlson**.
3. **Clinical variables.** *Add Variable* → search "essential hypertension",
   table `condition_occurrence`, format **Binary (0/1)**. *Add Variable* →
   search "hemoglobin A1c", table `measurement`, format **Mean**.
4. **Labels.** *Plan Options* → tick **Translate concept IDs to labels**.
5. **Output.** *Add Output* → name `cohort`, type **Wide**.
6. **Run.** Click ▶. The Execution Results card reports the `cohort` table is
   assigned on each server with its dimensions and columns.
7. **Analyse** the server-side result:

```{r analyse}
# The Run button assigned the result as a server-side symbol (e.g. D_cohort).
# Continue with ordinary DataSHIELD analysis on it:
dsBaseClient::ds.summary("D_cohort$age")
dsBaseClient::ds.glm("hypertension ~ age + charlson + hba1c_mean",
                     data = "D_cohort", family = "binomial")
```

Prefer code? Copy the **Generated Code** card instead of clicking Run — it is
the exact `omop_recipe()` / `recipe_execute()` script for the same extraction.

## What is *not* in the builder (use the R API)

A few advanced capabilities are intentionally left to the programmatic API for
now:

- **`value_bin`** filter — it needs a live `ds.omop.safe.cutpoints()`
  round-trip, which doesn't fit the build-then-run flow.
- **Full `omop.temporal`** (`index_window`, `calendar`, `min_gap`) — only the
  common *first/last N events* selection is surfaced in the UI.

Both remain available from R; see the *Extraction* and *Cohorts* vignettes.

## Cheat sheet

- **Who** → Add Filter (12 types) + Group Filters (AND/OR).
- **What** → Add Variable (21 formats, + descendants) · Block · Derived (scores).
- **Shape** → Add Output (wide / long / survival / …, + date handling).
- **Readable categories** → Plan Options ▸ *Translate concept IDs to labels*.
- **Run** → ▶, then read dimensions/columns and analyse the server-side symbol
  with `ds.*`. Row-level data never leaves the node.
