---
title: "Data Extraction with the Plan DSL"
description: "Build typed longitudinal extraction plans and output representations."
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Data Extraction with the Plan DSL}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(eval = FALSE)
```

## Overview

The **extraction plan DSL** specifies a typed subset of data and transformations
over the OMOP CDM. A plan is a serializable object describing a cohort, one or
more outputs, their representations, and global options. Plans are built
client-side, inspected against the server schema, and executed to produce
server-side `omop.table` objects. They are not raw SQL and do not permit
arbitrary joins or output schemas.

For higher-level, named, save/loadable extractions, the **recipe** API wraps the
plan (covered at the end of this vignette).

## Building a Plan

```{r basic}
library(dsOMOPClient)

plan <- ds.omop.plan()
plan <- ds.omop.plan.cohort(plan,
  spec = list(type = "condition", concept_set = c(201826)))
# Concept translation and federation-wide concept factors are ON by default
# (translate_concepts = TRUE, factor_concepts = TRUE). This example adds an
# inline index cohort because baseline/episode outputs need one.
```

### Baseline demographics

`ds.omop.plan.baseline()` adds a cohort-episode-level output of demographic
columns from the `person` table, plus optional derived columns (e.g.
`age_at_index`):

```{r baseline}
plan <- ds.omop.plan.baseline(plan,
  columns = c("gender_concept_id", "race_concept_id"),
  derived = c("age_at_index"),
  name = "demographics")
```

Baseline requires a cohort and emits one row per cohort episode. Raw birth
components are blocked; `age_at_index` is returned on the controller's public
age grid rather than as `year_of_birth` or exact completed age.

To merge reviewed one-row-per-person tables, use
`ds.omop.plan.person_level()`:

```{r person-level}
plan <- ds.omop.plan.person_level(plan,
  tables = list(
    person = c("gender_concept_id", "race_concept_id"),
    death = c("cause_concept_id")),
  name = "person_data")
```

Raw person-level sources are limited to `person` and `death`; repeatable tables
such as measurements, visits and observation periods must be requested as
events or reduced explicitly to features. The compiler will not pick one of
several records implicitly.

### Event-level data

`ds.omop.plan.events()` extracts rows from a clinical event table:

```{r events}
plan <- ds.omop.plan.events(plan,
  name = "conditions",
  table = "condition_occurrence",
  columns = c("person_id", "condition_concept_id", "condition_start_date"),
  concept_set = c(201826, 255573, 4329847),
  representation = list(format = "long"))
```

### Outcomes and feature sets

`ds.omop.plan.outcome()` is a convenience wrapper for a binary condition outcome
(it adds an event output with the `features` representation):

```{r outcome}
plan <- ds.omop.plan.outcome(plan, name = "copd", concept_set = c(255573))
```

`ds.omop.plan.features()` builds a person- or cohort-episode-level feature set
from explicit feature specs (see the `omop.feature.*` constructors). Episode
grain requires an index window and preserves `cohort_row_id`:

```{r features}
plan <- ds.omop.plan.features(plan,
  name = "hba1c_features",
  table = "measurement",
  specs = list(
    omop.feature.mean_value(concept_set = c(3004410)),
    omop.feature.latest_value(concept_set = c(3004410)),
    omop.feature.sd_value(concept_set = c(3004410))))
```

## Representations

Each event output uses one of four representations with an explicit row grain:

- **`long`** (default): one row per qualifying event or, with an index window,
  per event/episode match. Raw date columns are removed by default; request
  relative or binned handling explicitly.
- **`wide`**: pivots on the concept column at `grain = "person"` (default) or
  `grain = "episode"`. More than one event per grain/concept is rejected; use
  deterministic `event_select` or a feature reduction first. Federated wide
  output additionally requires a closed integer `concept_set` and
  `translate_concepts = FALSE`: every declared concept produces the same
  concept-ID-derived column on every node, including an all-`NA` column when
  absent locally, and undeclared concepts cannot enter the output.
- **`features`**: explicit reducers (counts, ever-flags, mean/sd, first/latest,
  durations, etc.) at person or episode grain. An `index_window` on wide or
  features requires `grain = "episode"`, preserving `cohort_row_id`.
- **`sparse`**: FeatureExtraction-style covariate triplets at person grain or,
  with an `index_window`, episode grain. Its complete `personRef` maps each
  `rowId` to the pseudonymized person and, for episode grain,
  `cohort_row_id`; a missing covariate row means implicit zero.

```{r repr}
plan <- ds.omop.plan.options(plan, translate_concepts = FALSE)
plan <- ds.omop.plan.events(plan, name = "labs_wide",
  table = "measurement", concept_set = c(3004410),
  temporal = omop.temporal(
    index_window = list(start = -365L, end = 0L),
    event_select = list(order = "last", n = 1L)),
  representation = list(format = "wide", grain = "episode"))
```

### Other longitudinal output contracts

The lower-level plan API also provides:

| Output | Grain and current limit |
|---|---|
| baseline / cohort membership | one row per `cohort_row_id` |
| survival | episode-grain survival, named competing risks, recurrent events, counting-process risk sets, or expanded multi-state transition risk sets; first/last/all event policies as allowed by the selected format; cohort, observation-period, death and administrative censoring |
| intervals-long | one source interval per cohort episode with start/end days from index |
| temporal covariates | sparse concept-by-time-bin values keyed by episode `rowId`, restricted to the unique OMOP observation period covering index, plus `personRef` |
| person-period | observation-aware episode-by-regular-bin roster plus sparse binary/count covariates; index-relative time only |
| sparse | person or indexed episode grain; complete `personRef`; no covariate row means implicit zero for a roster member |

Overlapping episodes can legitimately repeat one clinical event under different
`cohort_row_id` values. `event_select` is applied per episode when an index
window is present, and otherwise per person. It selects globally within that
grain by default. Set `event_select = list(order = "first", n = 1,
by = "concept")` for independent first/last-N selection per concept, or use
explicit `first_value`/`latest_value` feature reducers for concept-set-specific
values. `min_gap = 30L` chains same-concept events no more than 30 days apart;
the full policy can declare `list(days = 30L, by = "concept", keep = "first")`.
Collapse is per cohort episode when an index window is present and uses the
standard OMOP event primary key to break same-date ties deterministically.
Named person-level OHDSI estimands deliberately reduce recurrent index episodes
to one deterministic episode per person and pre-aggregate recurrent outcomes;
this prevents joins from multiplying person-time while still counting repeated
outcome events. General long, episode, recurrent-event, counting-process and
multi-state outputs do not perform that person-level reduction.
Longitudinal windows, bins, washout and risk intervals currently use calendar
days, consistent with required OMOP `*_date` fields and the federated date
contract. Optional `*_datetime` values do not change ordering or create sub-day
risk sets. The source primary key stays internal, so repeated attachments are
not linkable across episodes through a released high-cardinality event token.

A regular person-period panel is available through
`ds.omop.plan.person_period()` or Recipe output `type = "person_period"`. It
emits `personPeriods`, `temporalCovariates`, `covariateRef`, `timeRef`, and
`personRef`. Every cohort episode must have exactly one OMOP
`observation_period` covering its index date. Only bins intersecting that period
appear in `personPeriods`; `startDay`/`endDay` remain the requested bin limits,
while `observationStartDay`, `observationEndDay`, and inclusive `daysObserved`
describe the observed part. A missing sparse tuple means zero only when its
`(rowId, timeId)` roster row exists. The descriptive panel intentionally does
not treat cohort end or death as observation/censoring. Use survival or
counting-process output for explicit time-at-risk, washout, tie, and censoring
policies. Multi-table long recipes split into one output per table rather than
fabricating an arbitrary cross-table join.

### Multi-state disease courses

`format = "multi_state"` treats each named OMOP outcome as a state observation.
The public transition graph may be an `mstate`-style transition matrix, a named
adjacency list, or canonical `states`/`edges`. It can contain cycles and a person
may visit the same state repeatedly. Events that are not reachable from the
person's current state are skipped until a reachable event occurs; the graph is
never inferred from private site data.

```{r multistate, eval=FALSE}
transitions <- list(
  index = "disease",
  disease = c("recovery", "progression"),
  recovery = "disease",
  progression = character(0)
)

plan <- ds.omop.plan.survival(
  plan,
  outcomes = list(
    disease = list(table = "condition_occurrence", concept_set = 1L),
    recovery = list(table = "observation", concept_set = 2L),
    progression = list(table = "condition_occurrence", concept_set = 3L)
  ),
  format = "multi_state",
  transitions = transitions,
  initial_state = "index",
  state_hierarchy = c("progression", "disease", "recovery", "index"),
  tie_policy = "priority",
  name = "course"
)
```

In memory, the server assigns `<name>.msdata` in expanded `(Tstart, Tstop]`
form plus `<name>.transitionRef`; the former carries the usual `msdata`
transition-matrix attribute. Staged mode streams the ordered OMOP event source
through a bounded state machine directly to Parquet. Because Parquet does not
preserve R attributes, the transition dictionary is staged separately. Source
dates and event identifiers never enter either public component.

For simultaneous states, `priority` permits one reachable transition on the
day and `sequential` follows the public `state_hierarchy`. A data-dependent
`error` policy is deliberately unavailable because success versus failure would
reveal a cooccurrence bit outside DataSHIELD output controls. Sequential
ordering uses `state_step` only to place analytic times inside the same observed
calendar day; it does not claim clinical sub-day timing and never moves a
transition beyond censoring. Repeated records of the same state on the same day
collapse to one state observation because their order is not identifiable at
OMOP date resolution. Staged descriptors mark sequential time explicitly as
`calendar_day_with_public_within_day_offsets` rather than presenting those
analytic fractions as observed timestamps.

Malformed or incompletely observed longitudinal episodes are excluded by
server-side eligibility rules. The distinct-person gate is then applied to the
eligible risk set itself, so this filtering cannot expose a query-selective
data-quality error; only the ordinary generic DataSHIELD threshold can stop the
operation.

Representations that expand a request are subject to controller-configurable
operational caps. The defaults allow at most 1,000 explicit feature
specifications (`dsomop.max_feature_specs`), 1,000 concepts expanded into
wide/sparse/temporal output (`dsomop.max_pivot_concepts`), 5,000 final
wide/feature columns (`dsomop.max_output_columns`) and 10,000 temporal bins in a
window (`dsomop.max_temporal_bins`). Longitudinal first/last/nearest selection
retains at most 100 events per episode/source group
(`dsomop.max_events_per_group`). Filter trees are capped at depth 32, 1,024
nodes and 10,000 literal values, and a plan at 100 outputs. These are
controller-configurable memory/CPU safeguards rather than disclosure
guarantees. Across servers, construct a request within the lowest compatible
cap advertised by every participating site.

### Dates and ages

Use `omop.date_handling()` to make temporal precision explicit:

- `mode = "remove"` is the default and removes every date/datetime column;
- `mode = "relative"` emits integer days from the cohort index and replaces the
  index date itself with zero;
- `mode = "binned"` emits calendar year, month or Monday-based week bins;
- `mode = "absolute"` keeps raw dates only when the controller has enabled
  `dsomop.allow_absolute_dates`. That opt-in does not make raw dates harmless.

```{r dates}
plan <- ds.omop.plan.events(
  plan, name = "relative_conditions", table = "condition_occurrence",
  concept_set = c(201826),
  temporal = omop.temporal(index_window = list(start = -365L, end = 30L)),
  date_handling = omop.date_handling(mode = "relative", reference = "index")
)
```

When exact dates are disabled, selecting only some date columns for conversion
cannot leave the others untouched. Datetimes are converted using the
controller-owned timezone. Raw birth year/month/day are blocked. Derived age is
annual-resolution `reference_year - year_of_birth` and is released on the
controller's public age grid; it is not birthday-aware completed age. Closed
age/date filters must meet their configured minimum widths, and age-group labels
must be unions of the public grid. A federation negotiates the stricter common
contract described in the *Multi-Server* vignette.

## Typed Filtering and Scoping

### Row-level filter DSL

Beyond a concept set, `ds.omop.plan.events()` accepts a `filters` argument that
takes a **nested AND/OR tree of row-level predicates**. Build leaves with the
`omop_filter*` constructors and combine them with
`omop_filter_group(..., operator = "AND" | "OR")`:

```{r row-filters}
lab_bin <- ds.omop.safe.filter.value(
  table = "measurement",
  column = "value_as_number",
  threshold = 6.5,
  direction = "above",
  concept_id = 3004410,
  symbol = "omop",
  conns = conns
)

filt <- omop_filter_group(
  lab_bin,
  omop_filter_date_range(start = "2020-01-01", end = "2023-12-31"),
  operator = "AND")

plan <- ds.omop.plan.events(plan, name = "labs_filtered",
  table = "measurement", concept_set = c(3004410),
  filters = filt)
```

Disclosure note: the custom filter DSL is fail-closed. Categories use reviewed
membership/null predicates (`in`, `not_in`, `is_null`, `not_null`); dates use a
bounded `between` range that satisfies the controller's minimum width. Exact
matches and client-authored ordered thresholds (`==`, `!=`, `>=`, `>`, `<=`,
`<`) are **rejected server-side**. Numeric filtering uses only an authenticated
`value_bin` whose public grid was configured by the controller and issued by
the current resource session. Filtering on the person token or any protected /
identifier column is also rejected, and the distinct-person gate still runs on
the filtered result. Use `ds.omop.safe.filter.value()` for row filters or
`ds.omop.safe.filter.measurement()` for a population-level measurement range.

### Unit / type concept-column scoping

By default a table is profiled and filtered on its domain concept column. Pass
`concept_col` to scope to a different concept column instead -- e.g. a unit
(`unit_concept_id`), a record-type (`*_type_concept_id`), or a value concept
(`value_as_concept_id`):

```{r concept-col}
plan <- ds.omop.plan.events(plan, name = "labs_by_unit",
  table = "measurement", concept_set = c(8840),  # mg/dL
  concept_col = "unit_concept_id")
```

### Visit linkage

`visit_filter` restricts events to those occurring during a visit of a given
type, via an `EXISTS` join on `visit_occurrence_id`:

```{r visit}
plan <- ds.omop.plan.events(plan, name = "inpatient_dx",
  table = "condition_occurrence", concept_set = c(201826),
  visit_filter = list(concept_ids = c(9201)))  # inpatient
```

## Concept Translation

`translate_concepts` is **ON by default**. Concept-ID columns are replaced with
human-readable concept names from the vocabulary, so `condition_concept_id =
201826` is returned as `"Type 2 diabetes mellitus"`. Toggle it (and the related
options) with `ds.omop.plan.options()`:

```{r options}
plan <- ds.omop.plan.options(plan,
  translate_concepts = TRUE,
  factor_concepts = TRUE,
  block_sensitive = TRUE)
```

`factor_concepts = TRUE` (the default) coordinates releasable concept levels and
their ordering across servers in memory mode. This removes one common coding
mismatch, but does not by itself guarantee that an arbitrary downstream model
supports the output's grain or missing/suppressed levels. Recoding is
representation-only: every locally observed level must already pass the
server's person-support/cardinality gate and must be covered by the shared
contract. Otherwise the complete recoding call is a silent no-op. This keeps
the original values intact and prevents a partial factor contract from being
used as a category-presence probe; safe levels seen only at another site can
still be retained locally as empty factor levels.

## Plan Lifecycle

### Preview (server-side dry run)

`ds.omop.plan.preview()` sends the plan for a schema/structure preview that
returns per-server declared columns and a banded source-table support indicator
without creating assignment outputs:

```{r preview}
preview <- ds.omop.plan.preview(plan, symbol = "omop", conns = conns)

# Per-server, per-output details:
preview$server1$outputs$conditions$columns
preview$server1$outputs$conditions$n_persons_available
preview$server1$outputs$conditions$n_persons   # banded when available; else NA
preview$server1$outputs$conditions$n_persons_unavailable_reason
```

A person count is returned only for an honestly countable unscoped, unfiltered,
unreduced source. Cohort-, population-, filter-, temporal- and feature-scoped
outputs return `n_persons = NA`, `n_persons_available = FALSE` and a reason;
preview never substitutes the whole source table's population. An available
count is banded down to a multiple of `band_width` and suppressed below the
disclosure floor. Preview is structural and is not proof that execution will
pass its disclosure gate or DBMS-specific compiler. `ds.omop.plan.validate()`
calls the same limited endpoint; execution remains authoritative.

### Execute

`ds.omop.plan.execute()` runs the plan and assigns each output into the
DataSHIELD session under the symbols given in `out`. `out` is a **named**
character vector mapping each plan output name to a server-side symbol:

```{r execute}
ds.omop.plan.execute(plan,
  out = c(demographics = "D", conditions = "E"),
  symbol = "omop", conns = conns)

# D and E are protected server-side omop.table objects. Use a downstream
# method only after it has been reviewed for these classes and output grains.
dsBaseClient::ds.summary("D")
dsBaseClient::ds.dim("E")
```

With two or more servers, `validate`, `preview` and `execute` automatically
establish strict schema/semantic harmonization when the plan is not already
bound, or revalidate its existing contract. The contract binds the exact nodes,
required columns, compatible type families and, for vocabulary-dependent
plans, one identical non-missing reported vocabulary version. Calling
`ds.omop.plan.harmonize()` explicitly remains useful when the analyst wants to
inspect that contract before execution.

Before assignment, the client verifies that the exact namespace owned by the
requested outputs is fresh and refuses to start if any requested destination
already exists. Every expected component must then land on every node. A node
failure, incomplete composite output or factor-harmonization failure removes
and verifies all DataSHIELD-visible symbols created by the attempt. This is
symbol-level all-or-none cleanup; pre-existing symbols are never overwritten.

The server-side manipulation verbs use the same discipline: every source must
exist on every requested node, the destination must be fresh, and a partial
`merge`, `filter`, `select` or `bind_rows` assignment is removed and verified
across the federation. They never silently overwrite a pre-existing symbol.

For large extractions, `output_mode = "staged"` writes server-local files and
assigns descriptors inheriting from both `FlowerDatasetDescriptor` and the
package-neutral `OMOPStagedDatasetDescriptor`, instead of keeping the final
result as an R data frame. With `arrow` installed the files are Parquet
(otherwise the server uses its CSV fallback). Long event and interval outputs,
plus SQL-first temporal/person-period components, are fetched and sanitised in
bounded chunks into row groups of one Parquet file, so the complete result is
never materialised in the R session. Formats that require an R-side
transformation are materialised first and only then staged.

Staging is a local commit protocol: a pre-commit error removes the new dataset
directory, while a successful transaction writes the private `manifest.json`
version 2 only after every file and descriptor is complete. A v2 descriptor
binds its dataset/source/origin/token identity, path, format, owner-only
permissions and expiry. Its metadata also carries pseudonym-key ID/epoch, an
exact component `semantic_contract`, and an output-level `bundle_contract`
shared by sibling components.

The descriptor is intended for another reviewed server-side package or
application running under the same OS identity. Direct access by a different
service account is intentionally blocked by the 0700/0600 permissions, even on
a shared filesystem; such applications need a separately reviewed server-side
broker. It is not a client download URL or a general export capability. A
same-account consumer should call `dsOMOP::omopStagedDatasetPath()` instead of
trusting the embedded path. For sibling components, pass the first descriptor's
`bundle_contract`, non-secret pseudonym-key ID and epoch as the expected values;
compare an exact `semantic_contract` only between like-shaped datasets, because
`personRef`, `timeRef`, covariates and roster components intentionally have
different shapes. The resolver also validates identity, confinement, format,
expiry and file permissions. Consumers still need a reviewed/allowlisted
analysis contract; a descriptor class alone is not authorization. Person-
bearing staging requires a resource-scoped key provider; the deprecated global
legacy key is deliberately refused at this broader interoperability boundary.

The all-or-none guarantee above covers DataSHIELD-visible symbols, not a
distributed filesystem transaction. If another node fails after one node has
committed its local manifest, the client removes the descriptors but its
private files may remain registered with that server handle until handle
cleanup, disconnect or TTL cleanup.

The controller can bound each output's rows (`dsomop.max_staged_rows`), the
total bytes in one staged dataset directory (`dsomop.max_staged_bytes`), the
number of outputs in a staged plan (`dsomop.max_staged_outputs`), and retained
staging directories per OMOP handle (`dsomop.max_staging_dirs_per_handle`).
Directories and files are private to the server account (0700/0600), carry a
TTL, and are removed on handle cleanup; stale-TTL cleanup also runs when the
server package loads or attaches. It is not a background scheduler, so a
production deployment should additionally enforce filesystem quotas and run a
controller-owned periodic cleanup job.
`expires_at` is an upper-bound lease, not a persistence guarantee: explicit
handle/session cleanup can remove the directory sooner, and a consumer must
open the file promptly and handle an expired or missing descriptor.

## The Recipe API (named, reusable extractions)

A **recipe** is a higher-level, named container of populations, variables, and
outputs that compiles down to a plan. It is the most ergonomic path and the one
that supports saving/loading.

A recipe is authored as a **single declarative `omop_recipe()` call**: pass the
whole extraction as one nested expression built from the leaf constructors
(`omop_variable*`, `omop_filter*`, `omop_output`, `omop_population`,
`omop_variable_block`). Each argument (`variables`, `filters`, `outputs`,
`populations`, `blocks`, `cohort`, `options`) accepts a single object or a list
of objects, and the recipe is assembled in dependency order so later items can
reference earlier ones.

```{r recipe-build}
rec <- omop_recipe(
  populations = omop_population(
    id = "base",
    index_event = omop_index_event(
      table = "condition_occurrence",
      concept_id = 201826,
      primary_limit = "all"
    )
  ),
  variables = list(
    omop_variable(table = "person", column = "gender_concept_id",
                  format = "sex_mf"),
    omop_variable(table = "measurement", concept_id = 3004410,
                  format = "mean",
                  # per-variable row filters and a per-variable time window
                  filters = list(omop_filter_date_range(start = "2020-01-01",
                                                        end = "2023-12-31")),
                  time_window = list(start = -365, end = 0))
  ),
  outputs = omop_output(
    name = "study",
    type = "features",
    options = list(grain = "episode")
  )
)
```

Per-variable `filters` are forwarded as row-level predicates and `time_window`
(in days, relative to the cohort index date) becomes the variable's temporal
window when the recipe is compiled to a plan. Temporal wide/features outputs
must declare episode grain; person-grain temporal reduction fails closed.

Grouping many concepts from one table, declaring sub-populations, pinning a base
cohort, and toggling plan options all flow through the same call:

```{r recipe-full}
rec2 <- omop_recipe(
  blocks = omop_variable_block(
    table = "condition_occurrence",
    concept_ids = c(201820, 320128),
    concept_names = c("Type 2 diabetes", "Essential hypertension"),
    format = "binary"),
  variables = omop_variable_age(),
  filters = list(adults = omop_filter_age(min = 18, max = 65)),
  outputs = omop_output(name = "study", type = "wide"),
  cohort = 42,                              # base cohort_definition_id
  options = list(translate_concepts = TRUE) # only supplied keys override defaults
)
```

A named `filters` list (e.g. `adults = ...`) uses each name as the filter ID.

### Validate, preview, execute

```{r recipe-run}
recipe_validate(rec, symbol = "omop", conns = conns)
recipe_preview(rec, symbol = "omop", conns = conns)

# Ergonomic execute: with no `out`, symbols are derived from the output names
# (here "study" -> "D_study"). This is the simplest single-output form.
recipe_execute(rec, symbol = "omop", conns = conns)

# Or assign explicit symbols, exactly like ds.omop.plan.execute():
recipe_execute(rec, out = c(study = "D"), symbol = "omop", conns = conns)
```

### Save and load (YAML / JSON)

Recipes round-trip to disk; the format is inferred from the file extension (or
set `format =` explicitly):

```{r recipe-io}
recipe_save(rec, "study.yaml")        # or "study.json"
rec2 <- recipe_load("study.yaml")

# Or work with strings instead of files:
yaml_txt <- recipe_export_yaml(rec)
rec3 <- recipe_import_yaml(yaml_txt)
```

You can also emit runnable R for the whole recipe with `recipe_to_code(rec)`.
