---
title: "OHDSI QueryLibrary Sticky Redesigns"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{OHDSI QueryLibrary Sticky Redesigns}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(dsOMOPClient)
```

The QueryLibrary integration distinguishes an upstream clinical question from
its published SQL. The SQL is training/reference material and is never executed
by the sticky redesign path. Instead, a typed Recipe or Plan prepares a
server-side person-local `omop.table`, and a pinned mapping fixes one of the seven
sticky primitives and its longitudinal reducer.

This is deliberately not a QueryLibrary SQL runner or an arbitrary join/query
gateway. An upstream ID selects only a reviewed semantic mapping; it neither
submits the upstream SQL nor constructs a data-dependent query. Preparation and
release remain two explicit, typed steps.

There is also no `formal_dp` mode or certification flag. The client cannot
choose epsilon, seed, nonce, privacy epoch or rerolls. Actual releases use the
single server-owned sticky contract reported by `ds.omop.dp.status()`, currently
`sticky_person_bounded_noise_with_authenticated_lineage_and_nominal_accounting`.
That contract provides deterministic sticky replay, person-level contribution
bounds and durable nominal accounting; it must not be described as a general
formal-DP proof for unlimited distinct or semantically equivalent queries.

```{r catalog}
catalog <- omop_querylibrary_sticky_catalog()
table(catalog$statistic)
```

Every category domain, numeric/date break, bound, concept set, time window,
ordering rule and record cap must be public and fixed before data access.
Observed categories and data-dependent top-N selection are not valid inputs.

For example, PE12 can be redesigned as a coarse age-at-a-fixed-public-year
distribution. This avoids placing exact birth components in the prepared table
and uses public decade breaks:

```{r birth-year}
age_at_2025 <- omop_querylibrary_sticky(
  "PE12",
  variable = "age_at_2025",
  breaks = seq(0, 120, by = 10)
)
age_at_2025
```

Prepare one protected output explicitly and then release the mapping:

```{r execute, eval=FALSE}
recipe <- omop_recipe(
  variables = omop_variable_age(name = "age_at_2025", year = 2025),
  outputs = omop_output(name = "ages", type = "wide")
)
recipe_execute(recipe, out = c(ages = "D_ages"))

result <- ds.omop.querylibrary.sticky.release(
  "D_ages", age_at_2025, datasources = conns
)
```

Longitudinal numeric and categorical record histograms use a deterministic
declared order and a finite per-person record cap:

```{r records}
monthly_conditions <- omop_querylibrary_sticky(
  "CO19",
  variable = "month_number",
  breaks = seq(0.5, 12.5, by = 1),
  max_contributions = 12,
  order_by = "month_number"
)
```

Scalar record totals and site-local database-wide distinct drug cardinality are
also available only as explicitly capped estimands. The latter uses a complete
public vocabulary-derived concept domain and canonical server-owned ordering:

```{r bounded-counts, eval=FALSE}
condition_records <- omop_querylibrary_sticky(
  "CO10", max_contributions = 25
)

distinct_rxnorm <- omop_querylibrary_sticky(
  "DEX06",
  variable = "drug_concept_id",
  levels = public_standard_rxnorm_ids,
  max_contributions = 50
)
```

For one server, DEX06 estimates the distinct cardinality across that site's
database. Across multiple servers there is no privacy-safe cross-site entity
linkage for computing a set union: `pool = TRUE` therefore returns the explicit
sum of site-local bounded distinct counts and may count a concept once per site.
Use `pool = FALSE` when site-specific cardinalities are the intended estimand;
do not interpret the pooled sum as a global union cardinality.

Exact ZIP, source/free-text labels and patient/event rows remain unavailable.
They do not silently fall back to a different query:

```{r blocked, error=TRUE}
omop_querylibrary_sticky("PE08")
```

Use `omop_querylibrary_sticky_catalog(include_unavailable = TRUE)` to inspect
the complete 201-ID partition: 129 executable bounded redesigns, 54
vocabulary/reference metadata questions, and 18 blocked result shapes.
